# How the estimate is built {#sec-approach}

```{r}
#| include: false
source(here::here("chapters", "_common.R"))
```

## The arithmetic

Two quantities are estimated, one step apart.

**Potential revenue** is what the tax is owed on:

$$\text{Potential revenue} = 0.005 \times B_c \times \beta$$

- **$B_c$, the city's taxable sales.** Everything taxable that is sold and sourced *into* the City of Albany in a year, under the same rules the county uses, as it would appear in the Tax Department's taxable-sales statistics. The county's figure is published; the city's is not. Estimating it is the subject of @sec-apportionment and @sec-calibration, and it is the whole problem.
- **$\beta$ (beta), the shopping response.** The fraction of the base that stays put when the rate inside the city is half a point above the rate outside. It is taken as 1.0 (no loss), with lower values shown as sensitivities in @sec-sensitivity.

**Anticipated cash collections** are what the city would actually be paid:

$$\text{Anticipated cash} = \text{Potential revenue} \times (1 - \phi)$$

- **$\phi$ (phi), the reconciliation residual** (the project plan called it an administrative deduction). The gap between 4% of the published taxable sales and the cash the county actually receives. Refunds and the vendor filing credit reduce cash; tax collected at the DMV on private car sales, audit recoveries and late returns add to it; and the one-month offset between the sales tax year and the state fiscal year, together with the cells the Tax Department withholds from the published base, moves it either way. Its year-to-year swing is mostly timing and suppression rather than administration. It is measured below from Albany County's own figures and is small and slightly negative, so cash runs a little *above* potential revenue.

Every revenue figure on this site is anticipated cash. One point of bookkeeping matters for how the two steps combine. The apportionment in @sec-apportionment produces $B_c$, a taxable-sales figure, and $(1 - \phi)$ is applied to it. The comparison with other cities in @sec-calibration is fitted to those cities' *collections divided by their rate*, which is cash already, so what it produces for Albany is $B_c \times (1 - \phi)$ directly, and applying $(1 - \phi)$ a second time would count the residual twice. So the calibrated readings carry the residual inside them (the calibration cities' own residual, which the county test in @sec-calibration shows is within a few percent of Albany's), and the apportioned reading has it applied explicitly.

## Why the city's taxable sales are hard to know

The Tax Department publishes taxable sales by *taxing jurisdiction*: the state, New York City, the transit district and the 57 other counties, plus each city that has its own tax. Albany County is one; the City of Albany, having no tax, is not. There is no city row in the data, and no tabulation by ZIP code or by vendor address. The base has to be estimated.

One fragment of the city's base *is* observable. The Albany City School District has taxed utilities and telephone service inside the city at 3% since 1990, and its collections are published. Dividing them by 0.03 gives the city's utility-and-telephone base. It turns out to be `r fp(R$a_util)` of the county's, and it is the only measured city-versus-county share of any part of the base. @sec-checks uses it as a check, with a caveat about residential energy.

## The approach in six steps

1. **Start from the county's taxable sales by industry.** The Tax Department publishes them for `r nrow(appo)` industry groups. The three most recent complete sales tax years are averaged (March 2022 through February 2025), giving \$`r fb(B_k)` billion a year.

2. **Sort the industry groups into five classes according to where the tax law sources the sale.** Sales over the counter (stores, restaurants, hotels, personal services) are sourced to the store. Taxable goods and services bought by businesses (cleaning and security, software, repairs, equipment) are sourced to where the buying business takes delivery. Cars are sourced to where the *buyer lives*. Delivered goods (furniture, appliances, building materials) are sourced to the delivery address. Utilities and telephone service are sourced to where they are used. Governments are exempt purchasers, so their purchases are not in the base at all.

3. **For each class, measure the city's share of the county with the public data that best matches that rule.** Where the stores are, from the 2022 Economic Census: the city has `r fp(R$a_ecret)` of the county's retail receipts. Where the private-sector jobs are, from Census job-location data: the city has `r fp(R$a_wxp)` of the county's non-government jobs. Where car owners live, from DMV registrations: `r fp(R$a_mv)`. Where household income is, from the American Community Survey: `r fp(R$a_inc)`. These "sharing rules" are what the technical documents call *allocators*.

4. **Multiply each group's county sales by the city's share for its class, and add up.** This gives a raw city base of \$`r fb(B_pred)` billion, `r fp(B_pred / B_k)` of the county's total.

5. **Compare the method with cities whose collections are known.** Eighteen New York cities already levy their own sales tax, so a second estimate of their base can be inferred from their collections; seventeen reconcile with their county's figures. That inference has its own weaknesses, examined in @sec-calibration. Running the identical procedure on those cities shows that their collections run about 10% below the base the procedure gives, for reasons worked out there. Two cities are set aside for documented reasons that do not apply to Albany. Albany's apportioned figure, reduced by that typical gap, is **\$`r fb(est_pref$B_c_central)` billion**, and the scatter across the remaining cities sets the error range. The comparison is examined closely and given limited weight; @sec-results reports the estimate with and without it.

6. **Multiply by 0.5%** and allow for shoppers crossing the line: about **\$`r fm(REV)` million a year**, plausibly \$`r R$PLAN_LO`–`r R$PLAN_HI` million.

## The reconciliation residual, measured

Albany County's own data give $\phi$ directly: 4% of the published county base against what the county actually received.

```{r}
R$phi_tab |>
  transmute(`State fiscal year` = fy, `County taxable sales, $ billion` = B / 1e9,
            `4% of that, $ million` = 0.04 * B / 1e6,
            `Cash distributed to the county, $ million` = C / 1e6,
            `Cash relative to 4% of base` = fp(-phi, 2)) |>
  kb(digits = c(0, 3, 1, 1, 0), align = "lrrrr")
```

The gap is under 2.5% in every year and slightly negative on average: **the county received about `r fp(-phi, 2)` more than 4% of its published base**. The additions (DMV-collected tax on private car sales, audit, late returns) outweigh the subtractions. Applied at its observed mean to the apportioned base, it changes that reading by \$`r fm(abs(rev_base(B_pred) - 0.005 * B_pred))` million. It is also the first of the validations in @sec-checks, and @sec-calibration extends the same test to every county.

## Three terms used throughout

- **R, the concentration ratio.** The city's share of the county's taxable sales divided by its share of the county's population. R = 1 means the city has exactly its per-person share of sales; R = 2 means sales are twice as concentrated in the city as people are. Albany's R comes out at `r f2(R_cal)`. Among the seventeen taxing cities it runs from `r f2(min(route_b$R))` to `r f2(max(route_b$R))`.
- **Sharing rule** (technically, *allocator*). The measure used to say what fraction of a class of county sales belongs to the city: for example, the city's share of the county's retail receipts, or of its private-sector jobs.
- **Calibration.** Running the whole method on cities that already tax, comparing the result with the base their collections imply, and using the typical gap to adjust Albany's estimate and the scatter to size its error range.

## Three different "years"

Three calendars appear in the data, and they are easy to confuse.

| Calendar | Runs | Used for |
|---|---|---|
| Sales tax year | 1 March to end of February | The Tax Department's taxable-sales tables. "2024–25" is March 2024 to February 2025. |
| State fiscal year | 1 April to 31 March | The Tax Department's distributions to localities. "FY2025" ends 31 March 2025. |
| City and county fiscal years | Calendar year | Budgets. |

Sales tax year 2024–25 and state fiscal year 2025 overlap for eleven of twelve months, so they are matched to each other throughout. The three-year averages in this site cover sales tax years 2022–23 through 2024–25 and state fiscal years 2023 through 2025.
