# Checking the method where the answer is known {#sec-calibration}

```{r}
#| include: false
source(here::here("chapters", "_common.R"))
cv <- R$cv; loo <- R$loo; cal_pref <- R$cal_pref
```

The division of the county's sales in @sec-apportionment is a construction. It cannot be checked against Albany, whose base is unknown. But seventeen New York cities outside New York City impose their own sales tax, and for them a second estimate of the base exists: the city's collections divided by its rate. Running the identical construction on those cities and comparing the two estimates is what this chapter calls **calibration**.

Neither estimate is the truth. The apportioned base can be wrong in the ways @sec-apportionment describes. The collections-based base can be wrong too: the rate could be misread or in effect for only part of a year; the city's base could differ in definition from the county's; collections rise and fall with audits, refunds and late payments; and distributions to a city can be wrong at first and corrected later. So the chapter first asks how far the collections-based estimate can be trusted, then compares the two, then works through why they differ. Where the two disagree, this site does not assume the collections-based figure is right; it treats the gap as something to explain. For a city that would *receive* collections, though, the gap is what matters in the end.

## Inferring each city's base from its collections {#sec-inferring}

For a city that preempts part of its county's rate, the Tax Department publishes the cash distributed to the city each year. Dividing by the city's rate gives an estimate of the city's taxable sales. The rates come from Publications 718 and 718-A, checked against the Comptroller's 2020 table, and for two cities against the city code or the county treasurer.

### How good is the inference where it can be tested?

At the county level both quantities are published: the county's distributions and its reported taxable sales. So the same inference, collections divided by rate, can be tested on all 57 counties outside New York City. Where a county contains a taxing city, the amount the city preempted is added back to the county's collections.

```{r}
#| label: tbl-countytest
#| tbl-cap: "Collections divided by the county rate, as a ratio to reported taxable sales, for the 57 counties outside New York City, state fiscal years 2023–2025. Rates from Publication 718."
cs <- R$county_test_summary
tibble(Measure = c("County-years", "Median ratio", "Middle 90% of county-years", "Share within 5% of 1", "Share within 10% of 1",
                   "Typical year-to-year swing within a county (median)", "Largest year-to-year swing"),
       Value = c(fn(cs$n_county_years), f3(cs$median), sprintf("%s – %s", f3(cs$q05), f3(cs$q95)), fp(cs$within5, 0), fp(cs$within10, 0),
                 formatC(100 * cs$median_swing, format = "f", digits = 1) |> paste0(" points"),
                 formatC(100 * cs$max_swing, format = "f", digits = 1) |> paste0(" points"))) |>
  kb(align = "lr")
```

The inference is close: the median county-year ratio is `r f3(cs$median)`, `r fp(cs$within5, 0)` of county-years are within 5% of 1, and all are within 10%. Audits, refunds, late payments and timing move a county's collections by a few percent from year to year, not more. The counties furthest below 1 are the ones whose county tax on residential energy is at a *lower* rate than their general rate (Suffolk, Franklin, Tioga; @sec-resid-energy), which lowers collections relative to the base; Niagara, at `r f3(R$county_test_by_county$ratio[R$county_test_by_county$county == "NIAGARA"])`, is not explained. The full county table is in the technical appendix.

This test cannot see two things that matter for cities. It cannot see whether vendors report a sale under the city's code or the county's, because a sale mis-coded to the county still reconciles at the county level. And it cannot see whether a city's rate has been read correctly. Those are examined next.

### Is each city's inferred base stable?

If a city's inferred base were being moved around by audits, corrected distributions or a rate in force for part of a year, its share of the county base would jump from year to year. @tbl-stability shows the swing across the three years for each city, and whether every published source agrees on its rate.

```{r}
#| label: tbl-stability
#| tbl-cap: "Each city's inferred share of its county's taxable base: three-year mean, range, and swing (range as a share of the mean), with whether every published source agrees on the city's rate."
R$city_stability |>
  transmute(City = city, County = county, `Mean share` = fp(s_mean), `Lowest year` = fp(s_min), `Highest year` = fp(s_max),
            `Swing` = fp(swing, 0), `Rate agreed by every source` = ifelse(rate_all_agree, "yes", "no; resolved from footnotes, treasurer or city code")) |>
  kb(align = "llrrrrl")
```

Shares are stable. Only Ogdensburg, whose reporting code is new, swings more than 10%; most cities move by less than 6%, about what the counties do. So year-to-year noise in collections is not a large source of error here. Six cities have a rate the published sources disagree on (@sec-rates); the readings used are supported by two or three sources each, and the alternative reading is carried as a sensitivity.

### The rates


#### One unresolved point about rates {#sec-rates}

For six of the seventeen cities the published sources do not agree on the city's rate. Publication 718-A's percentage column reads 3% for Gloversville, Johnstown, New Rochelle, Norwich, Ogdensburg and Saratoga Springs; the Comptroller's 2020 table gives 2%, 2%, 2.5%, 1.5%, (absent) and 1.5%; and Publication 718-A's *own footnotes*, which state the county's rate inside each city, when subtracted from Publication 718's combined rate, reproduce the Comptroller's figures rather than the column. Ogdensburg was later verified at 3% from its city code and the Tax Department's notice ST-22-1; Saratoga Springs was confirmed at 1.5% by the county treasurer. This is not a claim that Publication 718-A is wrong; it is that its percentage column, read as the city's rate, contradicts its own footnotes and two other Tax Department-sourced documents. The rates on which three sources agree are used. @sec-sensitivity shows that the alternative reading changes the benchmark picture but barely moves Albany's estimate, because the calibration absorbs the shift.

## The two estimates, city by city

@tbl-calibcities shows, for each city, the county base from the taxable-sales table, the city base inferred from its collections, and the city base the apportionment gives. A ratio below 1 means the apportionment gives a larger base than the collections imply.

```{r}
#| label: tbl-calibcities
#| tbl-cap: "The apportionment run on every taxing city, beside the base inferred from its collections. County base: average of sales tax years 2022–23 to 2024–25. Collections: average of state fiscal years 2023–2025."
calib |>
  left_join(tibble(city = R$route_b$city, r_c = R$route_b$r_c), by = "city") |>
  transmute(City = city, County = county, `City rate %` = r_c,
            `County base, $ million` = B_k / 1e6,
            `City base from collections, $ million` = B_c_obs / 1e6,
            `City base from the method, $ million` = B_c_pred / 1e6,
            `Collections ÷ method` = ratio,
            `Population share` = fp(pop_share), `R (from collections)` = R) |>
  arrange(desc(`Collections ÷ method`)) |>
  kb(digits = c(0, 0, 1, 0, 0, 0, 3, 0, 2), align = "llrrrrrrr")
```

Fitting a straight line to the logarithms of the two city-base columns, with standard errors that allow for cities in the same county to move together, gives the calibration. A third fit, restricted to the `r length(R$TRUST_CITIES)` cities whose rate every source agrees on and whose inferred share is stable (`r paste(sort(R$TRUST_CITIES), collapse = ", ")`), is the fit that uses only the cities where the collections-based estimate is best supported.

```{r}
#| label: tbl-fits
#| tbl-cap: "Calibration fits. The slope says how the true base scales with the predicted one (1 = proportionally). σ is the typical miss, in logs. The last columns apply each fit to Albany."
bind_rows(cv |> transmute(Specification, n, `Slope` = beta, `se(slope)` = se, `σ (logs)` = sigma, `R²` = R2,
                          `Albany base, $ billion` = B_c / 1e9, `Revenue, $ million` = rev / 1e6),
          tibble(Specification = "Best-supported cities only (rate agreed by every source, stable share)", n = R$fit_trust$n,
                 Slope = R$fit_trust$beta, `se(slope)` = R$fit_trust$se_beta, `σ (logs)` = R$fit_trust$sigma, `R²` = R$fit_trust$r2,
                 `Albany base, $ billion` = R$est_trust$B_c_central / 1e9, `Revenue, $ million` = R$REV_trust / 1e6)) |>
  kb(digits = c(0, 0, 3, 3, 3, 3, 2, 1), align = "lrrrrrrr")
```

Across all seventeen cities the typical miss is σ = `r f3(fit$sigma)` in logs. Because the fit is multiplicative, a one-σ miss is **`r sd_up(fit$sigma)` on the high side and `r sd_dn(fit$sigma)` on the low side**; the two are not symmetric. That is well beyond the 0.25 the project plan set as a tripwire. Two of the seventeen misses have verifiable structural causes that do not apply to Albany, discussed below; setting those two aside brings σ to `r f3(fit_pref$sigma)`, a one-σ miss of `r sd_up(fit_pref$sigma)` / `r sd_dn(fit_pref$sigma)`, and that is the fit used. Both fits are reported throughout.

The calibration pulls Albany's apportioned base of \$`r fb(B_pred)` billion down to \$`r fb(est_pref$B_c_central)` billion (a `r fp(1 - est_pref$B_c_central / B_pred, 0)` reduction), because collections imply a smaller base than the apportionment for most cities: the typical ratio is `r f3(R$gm_pref)` among the fifteen and `r f3(R$gm_all)` across all seventeen. The fit on the best-supported cities alone gives \$`r fm(R$REV_trust)` million against the preferred fit's \$`r fm(REV)` million, so restricting to the cities where the collections-based estimate is most trustworthy does not change the answer much.

Whether the apportioned base or the collections-based base is nearer the truth is a fair question, and this site does not settle it. The county test above says collections measure the *reported* base closely. What it cannot say is how much of the city-level gap is real (sales that will never show up under a city's code) and how much is an artefact. For a city that will receive collections, the distinction is less important than it seems: what Albany would be paid is what shows up under its code. @sec-results therefore reports both figures, the apportioned \$`r fm(R$REV_raw)` million and the calibrated \$`r fm(REV)` million, and the planning range spans them.

## Why the method misses

The ratios in @tbl-calibcities are not noise. The city will receive **collections**, so what matters in the end is not the economic base but the base that shows up in returns under the city's code, and the reasons the two differ are worth understanding.

### The miss is a county phenomenon

Differences between counties explain **`r fp(R$cty_fe, 0)` of the variation** in the observed-to-predicted ratio. Cities in the same county miss in the same direction by similar amounts: both Fulton County cities come in above prediction, both Cattaraugus County cities far below, all four Westchester cities below 1. Whatever is going on is mostly a property of the county, not of how well the city's own sharing rules resolve. It is not, unfortunately, a predictable property: regressing the county-level ratio on the county's taxable sales per person, the obvious candidate, gives the expected sign but nowhere near significance on eleven counties.

### Four mechanisms

1. **Vendors code sales to the county, not the city.** A vendor selling into a taxing city must use that city's reporting code. Publication 718 warns that using ZIP codes for this "results in a high degree of inaccurate tax reporting." Errors default to the familiar county code, which understates the city's collections and overstates the county's, and the identity test *cannot detect it*, because the total still reconciles. The direct evidence is Ogdensburg, whose code was created on 1 March 2022: its implied share of the county base has climbed every year since (3.01%, 3.38%, 3.45%) while long-established Ithaca's sits flat. A new code takes time to propagate into vendors' systems. **Albany would be a new code too**, which is a first-year issue (@sec-timing).

2. **Capital-intensive activity outside the city that jobs cannot see.** The business rule uses jobs, but a data centre, a factory or a resort makes very large taxable purchases with few employees. St. Lawrence County reports \$88 million in data processing and hosting, essentially none of it in Ogdensburg. Warren County's taxable sales per person are twice the typical county's, with amusement parks and hotels in Lake George and Queensbury rather than Glens Falls. The jobs rule gives the city too much of these county bases. This is the same concern raised in @sec-business-rule.

3. **Economic Census receipts are gross, not taxable.** The store rule takes the city's share of *receipts*; what is being divided is *taxable* sales. Where the city's retail mix is more heavily exempt than the county's (groceries and pharmacy rather than restaurant meals and building materials), the receipts share overstates the taxable share. Auburn is the clearest case: 62% of Cayuga County's retail receipts, 35% of its population, and only 37.5% of its taxable base.

4. **Exempt institutions inside cities.** Hospitals and universities are concentrated in central cities and are largely exempt purchasers, but their jobs are counted in the business rule. This is the objection that removed government from the rule, and it is not fully solved.

All four push the same way, which is why the method over-predicts the typical city by about `r fp(1 - R$gm_all, 0)`.

### City by city

```{r}
#| label: tbl-citywhy
#| tbl-cap: "The main reason each city's collections differ from the method's prediction."
tibble(City = c("Ogdensburg","Salamanca","Glens Falls","Yonkers","Auburn","Olean",
                "New Rochelle","White Plains","Saratoga Springs","Utica","Mount Vernon",
                "Norwich","Oneida","Ithaca","Johnstown","Rome","Gloversville"),
  Reason = c(
    "Jurisdiction code created 1 March 2022; implied share still rising (3.01% to 3.45%). Canadian cross-border traffic, which the city depends on, fell sharply over the same years. County base also carries about $88 million of data-centre activity in Massena, outside the city.",
    "The whole city sits on the Seneca Nation's Allegany Territory (Census area 0080). Retail there is largely outside the state and local base under the 1842 Buffalo Creek Treaty and 20 NYCRR 529.9, while the Economic Census counts the receipts.",
    "Warren County's base is Lake George and Queensbury, not Glens Falls. The county treasurer reports 2024 sales tax of $12.1 million in Queensbury, $6.4 million in Bolton and $4.2 million in Lake George against $5.1 million in the city: a city share of about 13%, almost exactly the 12.6% observed and well below the 19.7% predicted.",
    "Dense residential city in a county whose retail and commerce sit in central and northern Westchester. Lowest R of any taxing city in the state.",
    "Holds 62% of Cayuga County's retail receipts but 35% of its population and only 37.5% of the taxable base: the clearest case of gross receipts overstating the taxable share. Anomalous wholesale activity (paper products) outside the city.",
    "Regional retail centre, but the county base includes the Ellicottville resort area and the Salamanca tribal complex, neither in Olean.",
    "Close to the norm; Westchester county-level effect.",
    "Close to the norm; a regional commercial centre, so the rules work.",
    "Close to the norm once the rate is right (1.5%, confirmed by the county treasurer). County base inflated by GlobalFoundries-area activity in Malta, outside the city.",
    "Close to the norm. Oneida County's base is split between Utica and Rome.",
    "Close to the norm.",
    "Close to the norm. Chenango County exempts clothing under $110 outside the city of Norwich only, so city and county bases differ in composition.",
    "Dominant commercial centre of a small rural county; the rules slightly understate that dominance.",
    "Students concentrated in the city generate retail that neither the residence nor the jobs rule sees; Cornell and Ithaca College are exempt purchasers but their students are not.",
    "Small city dominating a small county, as with Gloversville.",
    "Oneida County; the method splits the county between Rome and Utica more evenly than collections do.",
    "Fulton County is small and its two cities dominate its commerce; the suburb-weighted rules understate them.")) |>
  kb(align = "ll")
```

## Two cities set aside, and why that is defensible

Two of these reasons are verified from independent sources and demonstrably do not apply to Albany: Salamanca's tribal territory, and Ogdensburg's brand-new code combined with a border-traffic collapse. Dropping those two, and only those two, transforms the fit:

```{r}
#| label: tbl-twofits
#| tbl-cap: "The fit with and without the two structural exclusions."
bind_rows(
  tibble(Calibration = "All 17 cities", n = fit$n, Slope = fit$beta, `se(slope)` = fit$se_beta,
         `σ (logs)` = fit$sigma, `R²` = fit$r2, `Typical ratio` = R$gm_all,
         `Albany base, $ billion` = est$B_c_central / 1e9, `Revenue, $ million` = REV_all / 1e6,
         `68% range, $ million` = sprintf("%s – %s", fm(rev(est$lo68)), fm(rev(est$hi68)))),
  tibble(Calibration = "Excluding Salamanca and Ogdensburg (used)", n = fit_pref$n,
         Slope = fit_pref$beta, `se(slope)` = fit_pref$se_beta, `σ (logs)` = fit_pref$sigma,
         `R²` = fit_pref$r2, `Typical ratio` = R$gm_pref,
         `Albany base, $ billion` = est_pref$B_c_central / 1e9, `Revenue, $ million` = REV / 1e6,
         `68% range, $ million` = sprintf("%s – %s", fm(rev(est_pref$lo68)), fm(rev(est_pref$hi68))))) |>
  kb(digits = c(0, 0, 3, 3, 3, 3, 3, 2, 1, 0), align = "lrrrrrrrrr")
```

This is a judgment, and it is the one place on the site where dropping data changes the answer materially. The sequence should be stated plainly: the two cities were noticed *because* they were the largest misses, and the reasons were then investigated. What makes the exclusions defensible is not the order of discovery but that the reasons come from sources outside this analysis (the Census block files for Salamanca; the Tax Department's notice and the city code for Ogdensburg) and that neither transfers to Albany. Ogdensburg's case is the weaker of the two, because part of its evidence (the rising share) is a pattern in the same data.

The check a reader can audit is to refit dropping each city in turn and see which removals actually change the typical miss:

```{r}
#| label: tbl-loo
#| tbl-cap: "Leave-one-out: the fit with each city dropped in turn."
loo |> mutate(`Revenue, $ million` = rev(B_c) / 1e6) |>
  transmute(`City dropped` = dropped, n, Slope = beta, `σ (logs)` = sigma, `Revenue, $ million`) |>
  arrange(`σ (logs)`) |>
  kb(digits = c(0, 0, 3, 3, 1), align = "lrrrr")
```

Only the two excluded cities move σ appreciably; removing any other single city leaves it between `r f3(min(loo$sigma[!loo$dropped %in% c("(none)", names(R$CALIB_EXCLUDE))]))` and `r f3(max(loo$sigma))`. The exclusions are not one arbitrary pair among many. **Both fits are carried through everything that follows**, and a reader who rejects the exclusions should read \$`r fm(REV_all)` million with a 68% range of \$`r fm(rev(est$lo68))`–`r fm(rev(est$hi68))` million instead.

## The shopping response

The add-on makes the rate 8.5% inside the city against 8.0% outside, a 0.46% change in the tax-inclusive price at the city line. Vehicles are unaffected (sourced to residence). Business purchases, utilities and most services do not move. Only the over-the-counter classes, `r fp(R$store_share_of_base, 0)` of the base, are mobile at all, and only near the boundary. The central estimate assumes no loss. The research on cross-border shopping (Agrawal 2015; Baker, Johnson and Kueng 2021) establishes that responses exist and concentrate near rate boundaries, but neither study measures a half-point difference inside one county, so the lower cases in @sec-sensitivity are illustrative. Note the units: a 3% loss of base against a 6.25% rise in the tax is an elasticity of about 0.5 with respect to the *rate*, but against the tax-inclusive price, which rises only 0.46%, it is an elasticity of about 6.5, which would be a large response, not a modest one.
