# Alternative methods {#sec-alternatives}

```{r}
#| include: false
source(here::here("chapters", "_common.R"))
```

The industry-by-industry division was chosen because it is the only approach that respects how the tax law actually sources each kind of sale *and* can be tested against cities whose answer is known. But it is worth seeing what simpler and different methods give, because their spread is itself information.

```{r}
#| label: tbl-alts
#| tbl-cap: "Alternative ways of estimating the city's taxable sales, and the revenue each implies."
yh <- R$yonk_half; yp <- R$yonk_pop
shares <- c(pop, R$a_inc, R$a_hh, R$a_ecret, R$a_ecall, R$a_pay, R$a_wxp, R$a_work, R$a_util,
            median(route_b$R) * pop, median(R$pc_cities) * P_c / B_k, NA, NA, B_pred / B_k, s_cal)
alts <- tibble(
  Method = c("A. Population share of the county base",
             "B. Household income share (ACS)",
             "C. Household share (ACS)",
             "D. Retail receipts share (Economic Census)",
             "E. Receipts share, all matched sectors (Economic Census)",
             "F. Private payroll share (Economic Census)",
             "G. Private jobs share, excluding government (LODES)",
             "G'. All jobs share, including government (LODES)",
             "H. Measured utility-tax share applied to the whole base",
             "I. Typical taxing city: median R of the 17 cities × Albany's population share",
             "J. Taxable sales per person in the 17 cities (median) × Albany's population",
             "K. Yonkers analogy: what its half point raises per person × Albany's population",
             "K'. Yonkers analogy, adjusted for Albany's higher concentration ratio",
             "L. Industry-by-industry division, before calibration",
             "M. Industry-by-industry division, calibrated (this book)"),
  `City share` = ifelse(is.na(shares), "", fp(shares)),
  `City base, $ billion` = shares * B_k / 1e9,
  `Revenue, $ million` = rev(shares * B_k) / 1e6,
  `Likely direction of error` = c(
    "Ignores structure entirely; a coincidence that it lands near the centre",
    "Low: treats business purchases and commuter and visitor spending as if made where people live",
    "Low, for the same reason; Albany households are small",
    "Low: retail only, ignores services and business purchases, which are more concentrated in the city",
    "High: receipts at the seller's location; treats consumer retail as concentrated as offices and hospitals",
    "High: payroll concentrates in high-wage office and health jobs; omits government",
    "High: treats the entire base, including retail and vehicles, as if bought where private employers are",
    "Higher still: adds government jobs, whose employers are exempt purchasers",
    "High: utilities are the most city-concentrated component of any base, and the measure includes residential energy",
    "Assumes Albany is a typical taxing city; the 17 span a fourfold range",
    "Low: the 17 are mostly small cities with far less commercial and government activity per resident",
    "Low: Yonkers has the lowest R of any taxing city; its per-person base is not Albany's",
    "Circular (uses this book's R), but shows the Yonkers add-on scales consistently",
    "High relative to observed cities: over-predicts the typical calibration city by 10–20%",
    "—"))
alts$`Revenue, $ million`[12] <- yh / yp * P_c / 1e6
alts$`Revenue, $ million`[13] <- yh / yp * P_c / 1e6 * R_cal / R$R_yonk
alts |> kb(digits = c(0, 0, 2, 1, 0), align = "lrrrl")
```

Reading the table:

- The **residence-side** methods (B, C, D, J) cluster low; the **workplace-side** methods (E, F, G, G', H) cluster high. Each side is right about part of the base and wrong about the rest. The industry-by-industry division is the weighted combination and lands between them, which is the strongest simple argument that it is doing what it should.
- The **typical-taxing-city** transfer (I) gives \$`r fm(rev(median(route_b$R) * pop * B_k))` million, but "typical" spans a factor of four across the seventeen cities, so it is a weak anchor.
- The **Yonkers analogy** (K) is the only method that uses the yield of an actual half-point add-on. Yonkers' special rate is 1.5% (1% plus the 0.5%), so a third of its special-tax collections, \$`r fm(yh)` million for `r fn(yp)` people, is what half a point raises there: \$`r formatC(yh / yp, format = "f", digits = 0)` per resident, or \$`r fm(yh / yp * P_c)` million scaled to Albany's population. That is low because Yonkers is the least commercially concentrated taxing city in the state (R = `r f2(R$R_yonk)`); adjusting for Albany's higher R gives \$`r fm(yh / yp * P_c * R_cal / R$R_yonk)` million. The adjustment is circular, but the consistency is reassuring in one respect: a half-point difference at a city boundary has been administered by the Tax Department for years.
- The **uncalibrated division** (L) is \$`r fm(rev(B_pred))` million. The calibration's downward correction is the single largest methodological choice after the class assignments themselves, and it is worth `r fp(1 - REV / rev(B_pred), 0)`.

## Methods not implemented, and why

- **A Tax Department special tabulation.** The Office of Tax Policy Analysis holds return-level data with vendor location and could tabulate taxable sales inside the city limits directly, as it does for counties considering rate changes. This would replace everything in this book with a measurement. **It is the single most valuable next step**, and the city should request it; the school-district utility tax already gives the Department a precedent for accounting separately for sales within the city of Albany.
- **A consumer-spending build-up**: resident income times taxable-spending propensity, plus commuter and visitor inflows. Data-hungry, and the inflow term is exactly the unobserved quantity; it would not be independent of the assumptions already made.
- **A retail catchment model.** Would refine the store rule but not the business rule, which is where the uncertainty is.
- **Taxable commercial assessed value** from the city and county assessment rolls as an alternative business sharing rule. Feasible and cheap; not pulled here. It would be a useful third opinion alongside jobs and business locations, and it leaves out exempt property automatically (@sec-limitations).
