# Checks, the preemption comparison, and timing {#sec-checks}

```{r}
#| include: false
source(here::here("chapters", "_common.R"))
```

## Validations

| # | Check | Result |
|---|---|---|
| 1 | 4% of the county base against the county's distributions, each year | Cash within `r fp(min(-R$phi_tab$phi))` to `r fp(max(-R$phi_tab$phi))` of 4% of base; passes |
| 2 | Economic Census: sum of Albany County places against the county total, retail | Places cover 99.2%; residual 0.8% |
| 3 | Utilities: the jobs rule against the directly measured school-district base | Rule `r fp(R$a_wxp)` vs measured `r fp(R$a_util)` (an upper bound; see below). Using the measured share changes revenue by \$`r fm(abs(R$REV_util_meas - REV))` million |
| 4 | Motor vehicles: DMV resident share against household-income share | `r fp(R$a_mv)` vs `r fp(R$a_inc)`; DMV lower, consistent with lower urban vehicle ownership |
| 5 | Calibration (share form) | Slope `r f3(fit_pref$beta)` (se `r f3(fit_pref$se_beta)`), not distinguishable from 1; typical miss `r f3(fit_pref$sigma)` on the preferred sample, `r f3(fit$sigma)` on all seventeen |
| 5a | Collections ÷ rate against reported taxable sales, all 57 counties | Median `r f3(R$county_test_summary$median)`; `r fp(R$county_test_summary$within5, 0)` of county-years within 5% |
| 6 | Preemption identity across the eleven calibration counties | County collections plus preempted amounts equal the county rate on the whole base to within 0.5–2.7% |
| 7 | Albany's concentration ratio against the seventeen observed cities | `r f2(R_cal)`, within `r f2(min(route_b$R))`–`r f2(max(route_b$R))`, near the median `r f2(median(route_b$R))` |

**Has the city's store share drifted since the 2022 Economic Census?** County Business Patterns counts establishments by ZIP code every year; splitting the ZIPs to the city with the same block-population weights used for vehicles gives a rough annual series for the city's share of the county's retail and food-service establishments:

```{r}
#| label: tbl-drift
#| tbl-cap: "City of Albany's share of Albany County retail (NAICS 44-45) and food-service (72) establishments, from County Business Patterns ZIP-code counts split to the city by block population."
R$cbp_drift |> transmute(Year = year, Sector = ifelse(sector == "44-45", "Retail trade", "Accommodation and food services"),
                         `County establishments` = county_estab, `City share` = fp(city_share)) |>
  arrange(Sector, Year) |> kb(digits = 0, align = "rlrr")
```

Employment is withheld at ZIP level for these sectors, so only establishment counts can be followed, and the ZIP split is approximate; but a marked post-2022 fall in the city's share would show here, and it does not.

Validation 3 is the only place a sharing rule can be tested against a direct observation of an Albany city base. The measured share (`r fp(R$a_util)`) sits above the private-jobs rule (`r fp(R$a_wxp)`) and close to the all-jobs figure (`r fp(R$a_work)`). That is not evidence for counting government jobs: the State's own utility purchases are exempt and appear in neither the school-district collections nor the county's taxable sales. Nor is the comparison clean. The school district taxes residential gas and electricity at its full 3%, while Albany County does not tax residential energy at all and, as @sec-resid-energy shows from the data, the county's utilities figure does not contain it. The numerator therefore includes a residential component the denominator lacks, and the measured share is an *upper bound* of the comparable figure; the cross-county comparison suggests residential energy is on the order of half of a taxing county's utilities base, so the bound may be loose. What it does suggest is that taxable commercial load, which concentrates where the office buildings are, is at least as city-concentrated as the jobs rule says, which is one reason the alternatives in @sec-apportionment that lower the business share are read as a range rather than as corrections.

## The preemption comparison {#sec-preemption}

Albany County distributes 40% of its collections to cities and towns by census population (Comptroller 2020, Appendix C). The city's receipt under the status quo is therefore 0.40 × `r fp(pop)` × county collections. Three-year averages:

| Option | \$ million a year |
|---|---:|
| Status quo: 0.40 × population share × county collections of \$`r fm(R$C_k)` million | `r fm(R$status_quo)` |
| Preempt 1.5 points: 1.5% × city base | `r fm(0.015 * est_pref$B_c_central)` |
| **Add on 0.5 points (this site)** | **`r fm(REV)`** |

The status-quo figure can be checked against the City's own books. The City's Interim Financial Report for the fourth quarter of 2025 reports sales and use tax revenue of \$`r fm(R$city_report$actual_thousands * 1e3)` million for 2025, against a budget of \$`r fm(R$city_report$budget_thousands * 1e3)` million. The three-year figure above, \$`r fm(R$status_quo)` million, is consistent with it, which independently confirms the distribution arithmetic the comparison rests on.

Preempting at the standard 1.5 points would *lose* about \$`r fm(R$PREEMPT_LOSS)` million a year against the existing distribution. The reason is structural: preemption pays only if the city's share of the county base exceeds its population share by a factor of 0.40 × 4 ÷ 1.5 = 1.067, and Albany's R is `r f2(R_cal)`. A city has to be a retail concentrator for preemption to work, and Albany, with its county's retail on Wolf Road and at Crossgates, is not one. The add-on avoids the trade-off because it leaves the distribution untouched. The add-on would add \$`r fm(REV)` million on top of the \$`r fm(R$status_quo)` million the city already receives, an increase of `r fp(REV / R$status_quo, 0)`.

## Timing and first-year cash {#sec-timing}

Local rate changes take effect on 1 March, 1 June, 1 September or 1 December, with 90 days' notice (Tax Law §1210(d), §1211), and transitional rules apply to contracts that straddle the change (Tax Bulletin ST-895). Distributions lag liability by one to two months, with the first two months of each quarter paid as estimates.

**A new jurisdiction code collects less than its base implies at first.** Ogdensburg's implied share of its county base was `r fp(R$ramp$rel_to_latest[1], 0)` of its third-year level in year one and `r fp(R$ramp$rel_to_latest[2], 0)` in year two, and year three may not yet be the end of the climb (@sec-calibration). That is one city, so it can frame a discount but not fix it, and because Canadian cross-border traffic fell over the same years, its rising share understates how fast a code propagates, so the discount errs on the cautious side. **10–20% below steady state in the first full year, tapering in the second**, is the range consistent with it. Albany would be a new code. It would also be a code that vendors could mis-apply in the other direction, because the postal city "Albany" extends well into Colonie and Guilderland (@sec-calibration); the first year's collections could therefore include some suburban sales, and later corrections could run either way. Applied to the central estimate, the first *full* year is more like \$`r fm(REV * 0.80)`–`r fm(REV * 0.90)` million before the cash-lag effect below, and the two effects compound.

On a calendar fiscal year with an average 1.5-month cash lag, a **1 March** start yields roughly \$`r fm(REV * 8.5 / 12)` million of cash in the first year (ten months' liability, about 8.5 months' cash) and the full amount from the second. A **1 December** start yields essentially nothing in the first calendar year. If revenue in a particular budget year is the objective, the start date matters as much as most of the modelling choices in @sec-sensitivity.
