# Sensitivity analysis {#sec-sensitivity}

This appendix varies each assumption in turn. The results chapter gives the conclusions; the tables are here.

```{r}
#| include: false
source(here::here("chapters", "_common.R"))
sens <- R$sens; cv <- R$cv; af <- R$alloc_fits; forms <- R$forms
```

## Choices inside the division of the county base

Each row of @tbl-onechange changes one choice, re-runs the whole division of the county base for Albany, and applies the central calibration to the result.

```{r}
#| label: tbl-onechange
#| tbl-cap: "One choice changed at a time. Revenue is the calibrated central estimate; the central calibration is applied without re-fitting."
sens |> transmute(Variant, `Raw city base, $ billion` = B_pred_v / 1e9, `Calibrated city base, $ billion` = B_cal_v / 1e9,
                  `Revenue, $ million` = Rev / 1e6, `vs central` = fp(Rev / REV - 1)) |>
  kb(digits = c(0, 3, 3, 1, 0), align = "lrrrr")
```

The family of choices that moves the answer most is the sharing rule for sales to businesses; the store rule, tested the same way in @sec-store-rule, moves it less. @sec-apportionment takes the business rule apart and @tbl-busalts there gives the consistent test, in which the calibration is re-fitted under each rule so that Albany and the calibration cities are treated identically. Two points from it bear repeating here.

**Counting government jobs is not a plausible specification.** Government purchases are exempt, so using government employment to distribute taxable business purchases means using a measure of *exempt* activity to divide up *taxable* purchases. The row appears because it was the original plan's choice and because the difference (\$`r fm(abs(sens$Rev[2] - REV))` million) is worth knowing, not because the higher figure is a defensible alternative. The calibration cities cannot adjudicate: government is only 4–13% of jobs there, so their predictions barely move when it is removed, whereas it is `r fp(R$pub_pct, 0)` of Albany County's jobs with `r fp(R$a_pub, 0)` of them in the city.

**Hospitals and universities.** The same logic applies to them, and the job data separate education and health care, so a bound is available: dropping those sectors as well as government gives a city share of `r fp(R$a_exx)` and, with the calibration re-fitted, revenue of \$`r fm(af$rev[af$key == "lodes_ex_exempt"])` million. It is a bound rather than an estimate, because the two sectors also contain taxable private employers, but it says the remaining exempt-institution exposure is worth at most about \$`r fm(REV - af$rev[af$key == "lodes_ex_exempt"])` million.

Using the directly measured school-district share for the utilities class instead of the jobs rule gives \$`r fm(R$REV_util_meas)` million.

Everything else is immaterial. The delivered-goods split moves the answer by under 0.1%; the e-commerce assignment, which the 2022 industry codes make unresolvable, by under 1%. Both are immaterial for the same reason: the store share and the household-income share for those groups are within a point of each other.

## The calibration sample and the rate question

**If Publication 718-A's percentage column is the true rate** for the six cities discussed in @sec-rates, their recovered bases fall (by a third for the 2%-to-3% cities, by half for the 1.5%-to-3% ones). Re-running the calibration on that reading:

```{r}
#| label: tbl-rates-sens
#| tbl-cap: "The share-form calibration under the two readings of the six cities' rates, all seventeen cities."
tibble(Reading = c("Rates agreed by Pub 718-A footnotes, Pub 718 and the Comptroller (used)", "Pub 718-A percentage column"),
       `Benchmark median R` = c(median(route_b$R), median(R$rb_alt$R)),
       Slope = c(fit$beta, R$fit_alt$beta), `σ (logs)` = c(fit$sigma, R$fit_alt$sigma),
       `Albany base, $ billion` = c(est$B_c_central, R$est_alt$B_c_central) / 1e9,
       `Revenue, $ million` = c(REV_all, rev(R$est_alt$B_c_central)) / 1e6,
       `68% range, $ million` = c(sprintf("%s – %s", fm(rev(est$lo68)), fm(rev(est$hi68))),
                                  sprintf("%s – %s", fm(rev(R$est_alt$lo68)), fm(rev(R$est_alt$hi68))))) |>
  kb(digits = c(0, 2, 3, 3, 2, 1, 0), align = "lrrrrrr")
```

The benchmark picture changes a lot; Albany's estimate does not.

**Which cities, and what form of comparison.** @sec-fits sets out the two forms and @tbl-forms gives both on both samples. @tbl-samples repeats the exercise across calibration samples, in the share form used and in the level form.

```{r}
#| label: tbl-samples
#| tbl-cap: "Calibration sample × form. Share form is the one used; the level form is the lower reading."
cv |> transmute(Sample = Specification, n, `Share form: slope` = beta, `σ` = sigma, `Revenue, $ million` = rev / 1e6,
                `Level form: slope` = beta_level, `σ ` = sigma_level, `Revenue, $ million ` = rev_level / 1e6) |>
  kb(digits = c(0, 0, 3, 3, 1, 3, 3, 1), align = "lrrrrrrr")
```

In the share form the answer is steady across samples, from \$`r fm(min(cv$rev))` to \$`r fm(max(cv$rev))` million. The level form ranges more widely, from \$`r fm(min(cv$rev_level))` to \$`r fm(max(cv$rev_level))` million, and its slope moves with the sample: the four Westchester cities carry the size gradient, and dropping them from the full seventeen pushes the level slope above 1 and the estimate to \$`r fm(cv$rev_level[cv$Specification == "Excluding Westchester (4 cities)"])` million, while dropping them from the preferred fifteen leaves the slope at `r f3(cv$beta_level[cv$Specification == "Preferred 15 excluding Westchester (11 cities)"])` and the estimate at \$`r fm(cv$rev_level[cv$Specification == "Preferred 15 excluding Westchester (11 cities)"])` million. The share form is used because it does not carry a county-size gradient into Albany; the level form is the more cautious reading in the sense of giving less revenue, not in the sense of resting on fewer assumptions.

## Shopping response and reconciliation residual

```{r}
#| label: tbl-beta
#| tbl-cap: "Shopping response, and the reconciliation residual on the apportioned reading."
b <- est_pref$B_c_central
bind_rows(
  tibble(Assumption = "No shopping loss (central)", `Revenue, $ million` = rev(b) / 1e6),
  tibble(Assumption = "3% loss of the whole base", `Revenue, $ million` = rev(b, 0.97) / 1e6),
  tibble(Assumption = "3% loss of the over-the-counter classes only", `Revenue, $ million` = rev(b) * (1 - 0.03 * R$store_share_of_base) / 1e6),
  tibble(Assumption = "10% loss of the whole base (aggressive)", `Revenue, $ million` = rev(b, 0.90) / 1e6),
  tibble(Assumption = "10% loss of the over-the-counter classes only", `Revenue, $ million` = rev(b) * (1 - 0.10 * R$store_share_of_base) / 1e6),
  tibble(Assumption = paste0("Apportioned reading, wedge as observed (", fp(phi, 2), ")"), `Revenue, $ million` = rev_base(B_pred) / 1e6),
  tibble(Assumption = "Apportioned reading, wedge of zero", `Revenue, $ million` = rev_base(B_pred, ph = 0) / 1e6),
  tibble(Assumption = "Apportioned reading, wedge of +2% (an explicit charge)", `Revenue, $ million` = rev_base(B_pred, ph = 0.02) / 1e6)) |>
  kb(digits = 1, align = "lr")
```

Neither factor moves the estimate by more than about 10% even under aggressive assumptions; the uncertainty in the base is much larger. The wedge applies only to the apportioned reading; the calibrated readings are fitted to cash and already contain it.

## Base year and growth

```{r}
#| label: tbl-years
#| tbl-cap: !expr sprintf("The calibrated share (%s) applied to each year's county base.", fp(s_cal))
by <- R$by_year
bind_rows(by |> transmute(Base = sales_tax_year, `County base, $ billion` = B / 1e9,
                          `City base, $ billion` = s_cal * B / 1e9, `Revenue, $ million` = rev(s_cal * B) / 1e6),
          tibble(Base = "Three-year average used (2022–23 to 2024–25)", `County base, $ billion` = B_k / 1e9,
                 `City base, $ billion` = est_pref$B_c_central / 1e9, `Revenue, $ million` = REV / 1e6)) |>
  kb(digits = c(0, 3, 2, 1), align = "lrrr")
```
