---
execute:
  freeze: auto
---

# Technical report {#sec-technical}

This appendix is the original technical document, retained in full for readers who want the detail behind the chapters. It uses the technical vocabulary (allocator, apportionment, NAICS, σ) that the chapters translate; see the glossary.


```{r setup}
#| include: false
library(here)
source(here("R", "04_allocate.R"))
suppressPackageStartupMessages({library(gt); library(ggplot2)})

ALB <- list("Albany city, New York", "3601000", "36001", "ALBANY", "ALBANY")
POP_SHARE <- {
  p <- read_pop()
  p$place$P_c[p$place$place_fips == "3601000"] / p$county$P_k[p$county$county_fips == "36001"]
}

appo   <- do.call(apportion_city, ALB)
B_k    <- sum(appo$B_kg)
B_pred <- sum(appo$B_cg, na.rm = TRUE)

calib  <- build_calibration()
fit    <- fit_calibration(calib)
fit_nw <- fit_calibration(calib, drop_westchester = TRUE)
## Preferred: excludes the two cities with a verified structural reason that
## does not transfer to Albany (CALIB_EXCLUDE in R/05_calibrate.R; see the memo).
cal_pref <- calib |> filter(!city %in% names(CALIB_EXCLUDE))
fit_pref <- fit_calibration(cal_pref)
est_all  <- apply_calibration(fit, B_pred)
est      <- apply_calibration(fit_pref, B_pred)

phi_tab <- albany_phi()
phi     <- mean(phi_tab$phi)
BETA    <- 1.0
BETA_LO <- 0.97

rev_of <- function(b, beta = BETA) revenue_from_base(b, phi = phi, beta = beta)
REV     <- rev_of(est$B_c_central)

route_b <- summarise_route_b(build_route_b())
R_alb   <- (est$B_c_central / B_k) / POP_SHARE
R_raw   <- (B_pred / B_k) / POP_SHARE

C_k <- mean(read_distributions() |>
              filter(taxing_jurisdiction == "Albany County Sales and Use Tax",
                     fy %in% ANALYSIS_FYS) |> pull(amt))

M  <- function(x, d = 1) sprintf(paste0("%.", d, "f"), x / 1e6)
B  <- function(x, d = 2) sprintf(paste0("%.", d, "f"), x / 1e9)
PC <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
```

::: {.callout-note}
## The estimate

A 0.5 percentage-point City of Albany add-on sales and use tax, on the same base
Albany County taxes at 4%, destination-sourced, layered on the existing 8% with
no preemption and no change to county distributions, would raise approximately

### **\$`r M(REV)` million a year** {.unnumbered}

with a 68% range of **\$`r M(est$lo68 * 0.005 * (1-phi))`–`r M(est$hi68 * 0.005 * (1-phi))` million**
and a 90% range of **\$`r M(est$lo90 * 0.005 * (1-phi))`–`r M(est$hi90 * 0.005 * (1-phi))` million**,
in the price and activity levels of sales tax years `r min(ANALYSIS_FYS)-1`-`r max(ANALYSIS_FYS)`.

The band is wide, and honestly so: it is driven by how poorly the method
performs on the cities where the answer can be checked, not by sampling noise.
:::

## The identity

$$\text{Revenue} = 0.005 \times B_c \times (1 - \varphi) \times \beta$$

| term | meaning | value used |
|---|---|---|
| $B_c$ | City of Albany taxable sales and purchases | **\$`r B(est$B_c_central)` B** (calibrated) |
| $\varphi$ | administrative deduction, $1 - \frac{\text{county distributions}}{0.04 \times \text{county base}}$ | **`r PC(phi, 2)`** |
| $\beta$ | behavioral response to the 0.5-point differential | **`r sprintf("%.2f", BETA)`** central |

$B_c$ is published by nobody. Everything below is about estimating it and being
honest about how well that can be done.

## 1. The county base, and the reconciliation

Albany County's taxable sales and purchases come from the DTF quarterly file,
summed over all `r nrow(appo)` NAICS groups and averaged over the three sales
tax years ending `r max(ANALYSIS_FYS)`: **\$`r B(B_k)` billion**.

Validation 1 in §9 requires that 4% of that base reconcile to what the county
actually received.

```{r}
#| label: tbl-recon
#| tbl-cap: "Validation 1. County distributions against 4% of the published county base. The sales tax year runs March–February; the state fiscal year runs April–March, so the two are offset by one month."
phi_tab |>
  transmute(`Fiscal year` = fy,
            `County base ($B)` = B / 1e9,
            `4% of base ($M)` = 0.04 * B / 1e6,
            `Distributions ($M)` = C / 1e6,
            `Gap` = -phi) |>
  gt() |>
  fmt_number(`County base ($B)`, decimals = 3) |>
  fmt_number(c(`4% of base ($M)`, `Distributions ($M)`), decimals = 1) |>
  fmt_percent(Gap, decimals = 2) |>
  tab_options(table.font.size = pct(90))
```

The gap is under 2.5% in every year and averages `r PC(-phi, 2)` — the plan
called for "low single digits percent" and flagged anything over 10% as a
broken pull. It passes.

The gap is also *negative on average*: the county received slightly more than
4% of the published base. That is not an administrative surplus. It reflects the
one-month calendar offset, late and amended returns, use tax on purchases that
does not sit in the "taxable sales" measure, and the withheld cells discussed
below. The practical consequence is that **$\varphi$ is indistinguishable from
zero**, and it is carried at its observed mean of `r PC(phi, 2)` rather than
being invented.

::: {.callout-warning}
## Cell suppression

DTF withholds individual county × NAICS cells for disclosure, and `ny73-2j3u`
has no jurisdiction total row. Every county base here is therefore the sum of
its *published* groups. For Albany the withheld share is about 3.6% of groups;
the reconciliation above bounds their aggregate effect, since collections
reconcile to the published base within a couple of percent.
:::

## 2. Sourcing classes

Every one of the `r nrow(appo)` NAICS groups maps to exactly one class, recorded
with a rationale and a source in `data/crosswalk/naics_sourcing_class.csv`.

```{r}
#| label: tbl-classes
#| tbl-cap: "Sourcing classes and the allocator each uses. Base is the three-year mean county base."
appo |>
  group_by(sourcing_class) |>
  summarise(`Groups` = n(),
            `County base ($M)` = sum(B_kg) / 1e6,
            `Share of base` = sum(B_kg) / B_k,
            `Effective a_g` = sum(B_cg, na.rm = TRUE) / sum(B_kg),
            `City base ($M)` = sum(B_cg, na.rm = TRUE) / 1e6, .groups = "drop") |>
  mutate(Allocator = c(business = "LODES employment excl. public admin",
                       delivered_split = "50% Econ Census receipts / 50% ACS income",
                       motor_vehicle = "DMV resident registrations",
                       store = "Econ Census receipts",
                       utilities = "LODES employment excl. public admin")[sourcing_class],
         Class = sourcing_class) |>
  select(Class, Allocator, Groups, `County base ($M)`, `Share of base`,
         `Effective a_g`, `City base ($M)`) |>
  arrange(desc(`County base ($M)`)) |>
  gt() |>
  fmt_number(c(`County base ($M)`, `City base ($M)`), decimals = 1) |>
  fmt_percent(`Share of base`, decimals = 1) |>
  fmt_number(`Effective a_g`, decimals = 3) |>
  tab_options(table.font.size = pct(85))
```

Two choices deserve comment.

**Motor vehicles are 4411, 4412 and 9261.** Vehicle sales are sourced to the
purchaser's residence, not the dealer's location: Publication 838 instructs
dealers that "sales tax is collected at the combined rate in effect in the
local jurisdiction where the customer is a resident, regardless of where the
vehicle is delivered", and Form DTF-802 applies the same rule at DMV
registration. That reasoning does not extend to 4413 (parts, accessories and
tires), which is ordinary over-the-counter retail and is left store-based; the
project plan specified all of NAICS 441, and 4413 is \$65 M of a \$`r B(B_k)` B
base — it is in the sensitivity table. NAICS 9261, *Administration of Economic
Programs*, is added to this class: it carries \$8.5 B of statewide "taxable
sales" spread across counties in proportion to population and car ownership
(Albany County's share is exactly its population share), the signature of the
tax DMV collects at registration on private-party vehicle sales rather than of
state-agency vendor activity concentrated in Albany. This is inferred from the
pattern, not documented by DTF.

**There is no e-commerce class**, because NAICS 2022 eliminated sector 454 and
folded electronic shopping into the store categories. The plan assumed a
nonstore group to allocate by residence; no such group exists in the analysis
window and DTF publishes no split. This turns out not to matter much, and §7
shows why.

## 3. The allocators

```{r}
#| label: tbl-alloc
#| tbl-cap: "City share of the county under each allocator, with the population share for reference."
tibble(
  Allocator = c("LODES total employment",
                "  — of which public administration",
                "  — excluding public administration (used for business, utilities)",
                "Economic Census payroll, matched sectors",
                "Economic Census receipts, store classes (base-weighted)",
                "DMV resident registrations, model year ≥ 2024",
                "ACS aggregate household income (residence)",
                "ACS households",
                "Measured: school district utility tax ÷ 0.03",
                "Population (2020 Census)"),
  Share = c(work_share("3601000", "36001"),
            pubadmin_share("3601000", "36001"),
            work_share_ex_pubadmin("3601000", "36001"),
            ec_payroll_share("Albany city, New York", "001"),
            with(appo[appo$sourcing_class %in% c("store", "delivered_split"), ],
                 sum(a_store_used * B_kg) / sum(B_kg)),
            mv_share("3601000", "ALBANY"),
            resid_share("01000", "001", "agginc"),
            resid_share("01000", "001", "hh"),
            albany_measured_utility_share(),
            POP_SHARE)) |>
  gt() |>
  fmt_percent(Share, decimals = 1) |>
  tab_options(table.font.size = pct(88))
```

The spread across allocators is the whole story of this estimate. Albany city
holds **`r PC(work_share_ex_pubadmin("3601000","36001"))` of the county's
non-government jobs** (and `r PC(work_share("3601000","36001"))` of all jobs) but only
**`r PC(mv_share("3601000","ALBANY"))` of its vehicle registrations** — a
factor of two and a half, around a population share of `r PC(POP_SHARE)`. Which
allocator applies to which part of the base determines the answer.

Two results are worth pulling out.

**Government is exempt, so government jobs are left out.**
`r PC(pubadmin_share("3601000","36001"))` of Albany County's public-administration
jobs are inside the city, and public administration is
`r PC(with(lodes_place |> filter(cty=="36001"), sum(CNS20)/sum(C000)))` of all
county employment. The business class consists of taxable sales by
business-serving vendors sourced to the *customer's* location, so its allocator
has to stand for where purchasing businesses are; New York State and its
agencies are exempt purchasers (Tax Law §1116(a)(1)) and their jobs generate no
taxable purchasing of their own. What state workers spend at lunch and after
work is already in the store classes through the receipts of city restaurants
and shops. The central business allocator is therefore employment *excluding*
public administration; total employment is a sensitivity (§7), and the
calibration cannot distinguish the two because the calibration cities have so
little government employment.

**Validation 3.** The utilities class is allocated in the generic apportionment
by the business allocator, giving
`r PC(work_share_ex_pubadmin("3601000","36001"))`. It can also be *measured*: the
Albany City School District's 3% utility tax implies a city utility base of
`r PC(albany_measured_utility_share())` of the county's NAICS 22 + 517 base. The
measured share sits between the private-employment allocator and total
employment (`r PC(work_share("3601000","36001"))`) — and closer to total
employment, `r sprintf("%.1f", 100*abs(albany_measured_utility_share()-work_share("3601000","36001")))`
points from it against
`r sprintf("%.1f", 100*abs(albany_measured_utility_share()-work_share_ex_pubadmin("3601000","36001")))`
points from the allocator used. That is not evidence for counting government
jobs: the State's own utility purchases are exempt and appear in neither the
school-district collections nor the county's taxable sales, so the 48.9% is
*taxable* commercial and residential load, which concentrates where the office
buildings are rather than where the government payroll is. It does say the
business-purchase share of the city may sit a little above the private-employment
allocator, and it is the only place in this exercise where an allocator can be
checked against a direct observation of an Albany city base. Using the measured
share directly for that class is in the sensitivity table.

**Validation 4.** The DMV registration share
(`r PC(mv_share("3601000","ALBANY"))`) sits below the ACS income share
(`r PC(resid_share("01000","001","agginc"))`) and well below population. Albany
city households own fewer vehicles per person than the county average — an urban
core with students, transit and lower incomes. Registrations carrying
out-of-state ZIPs (Ft Lauderdale, Port Newark: leasing and fleet registrants,
`r PC(with(dmv_by_zip |> filter(county=="ALBANY"), sum(n[state!="NY"])/sum(n)))`
of the county total) have no in-county residence to assign and are dropped from
both numerator and denominator, which distributes them in proportion to the
resident pattern.

::: {.callout-note}
## ZIP codes were not assigned whole

Albany's ZIPs straddle Colonie, Guilderland and Menands. The ZCTA-to-place
relationship file publishes land area only, and area weighting badly understates
a dense central city: ZCTA 12203 is 40% of Albany city by area but
**62% by population**. Exact population weights were built instead from 2020
Census block populations joined to block→place (Block Assignment Files) and
block→ZCTA (the national ZCTA-to-block relationship file, stream-filtered to the
13 counties in this analysis rather than downloaded whole).
:::

## 4. The raw apportionment

$$B_c^{\text{pred}} = \sum_g B_{k,g} \times a_g = \textbf{\$`r B(B_pred)` billion}$$

which is `r PC(B_pred/B_k)` of the county base against a population share of
`r PC(POP_SHARE)` — an implied $R$ of `r sprintf("%.2f", R_raw)`.

```{r}
#| label: tbl-groups
#| tbl-cap: "The twenty largest NAICS groups by county base, showing the allocator resolved for each and the fallback flag."
appo |>
  arrange(desc(B_kg)) |>
  head(20) |>
  transmute(NAICS = naics_industry_group,
            Description = substr(description, 1, 44),
            Class = sourcing_class,
            `County base ($M)` = B_kg / 1e6,
            Allocator = allocator,
            `a_g` = a_g,
            `City base ($M)` = B_cg / 1e6,
            Fallback = ifelse(fallback, "yes", "")) |>
  gt() |>
  fmt_number(c(`County base ($M)`, `City base ($M)`), decimals = 1) |>
  fmt_number(`a_g`, decimals = 3) |>
  tab_options(table.font.size = pct(78))
```

Of the `r sum(appo$sourcing_class %in% c("store","delivered_split"))` groups
needing an Economic Census share,
`r sum(grepl("4-digit", appo$store_method[appo$sourcing_class %in% c("store","delivered_split")]))` resolved at the 4-digit level and
`r sum(grepl("3-digit", appo$store_method[appo$sourcing_class %in% c("store","delivered_split")]))` fell back to 3-digit. None needed
the sector level and none went unresolved.

## 5. Calibration

The apportionment above is a construction, not a measurement. The seventeen
cities from Phase 2 are the places where the same construction can be run and
then checked against a base that is actually observed.

```{r}
#| label: tbl-calib
#| tbl-cap: "The identical apportionment run on every calibration city, against the base implied by its own collections. Ratio below 1 means the method over-predicts."
calib |>
  transmute(City = city, County = county,
            `Observed ($M)` = B_c_obs / 1e6,
            `Predicted ($M)` = B_c_pred / 1e6,
            `Obs ÷ Pred` = ratio) |>
  arrange(`Obs ÷ Pred`) |>
  gt() |>
  fmt_number(c(`Observed ($M)`, `Predicted ($M)`), decimals = 0) |>
  fmt_number(`Obs ÷ Pred`, decimals = 3) |>
  data_color(columns = `Obs ÷ Pred`, palette = c("#d6604d", "#f7f7f7", "#4393c3"),
             domain = c(0.25, 1.25)) |>
  tab_options(table.font.size = pct(88))
```

Fitting $\log B_c^{\text{obs}} = \alpha + \beta \log B_c^{\text{pred}}$, with
standard errors clustered by county because Fulton, Cattaraugus, Oneida and
Westchester each contribute more than one city:

```{r}
#| label: tbl-fit
#| tbl-cap: "Calibration fits. Sigma is the residual standard deviation in logs and sets the width of the band."
mk <- function(lbl, f) tibble(Specification = lbl, n = f$n, Clusters = f$n_clusters,
                              `α` = f$alpha, `se(α)` = f$se_alpha,
                              `β` = f$beta, `se(β)` = f$se_beta,
                              `σ (logs)` = f$sigma, `R²` = f$r2)
c_ns  <- calib |> filter(city != "Salamanca")
f_ns  <- { d <- c_ns; m <- lm(log_obs ~ log_pred, d)
           ct <- lmtest::coeftest(m, vcov. = sandwich::vcovCL(m, cluster = d$county))
           list(n = nrow(d), n_clusters = n_distinct(d$county), alpha = coef(m)[1],
                se_alpha = ct[1,2], beta = coef(m)[2], se_beta = ct[2,2],
                sigma = summary(m)$sigma, r2 = summary(m)$r.squared) }
bind_rows(mk("All 17 cities", fit),
          mk("Excluding Salamanca and Ogdensburg (used)", fit_pref),
          mk("Excluding Westchester", fit_nw),
          mk("Excluding Salamanca", f_ns)) |>
  gt() |>
  fmt_number(c(`α`, `se(α)`, `β`, `se(β)`), decimals = 3) |>
  fmt_number(c(`σ (logs)`, `R²`), decimals = 3) |>
  tab_options(table.font.size = pct(85))
```

```{r}
#| label: fig-calib
#| fig-cap: "Observed against predicted city base, log scale. The dashed line is the fitted calibration, the grey line is equality. Albany's prediction is marked on the horizontal axis and its calibrated estimate on the vertical."
#| fig-height: 5.5
ggplot(calib, aes(B_c_pred / 1e6, B_c_obs / 1e6)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey75") +
  geom_abline(slope = fit$beta, intercept = fit$alpha - (fit$beta - 1) * log(1e6),
              linetype = "dashed", colour = "#b2182b") +
  geom_point(aes(colour = westchester), size = 2.8) +
  ggrepel::geom_text_repel(aes(label = city), size = 2.9, max.overlaps = 20, seed = 1) +
  annotate("point", x = B_pred / 1e6, y = est$B_c_central / 1e6,
           shape = 18, size = 5.5, colour = "#1b7837") +
  annotate("text", x = B_pred / 1e6, y = est$B_c_central / 1e6,
           label = "  Albany (calibrated)", hjust = 0, size = 3.2, colour = "#1b7837") +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c("FALSE" = "#4d4d4d", "TRUE" = "#762a83"),
                      labels = c("Rest of state", "Westchester"), name = NULL) +
  labs(x = "Predicted city base, $M (log scale)",
       y = "Observed city base, $M (log scale)") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

### Reading the fit

The slope is **`r sprintf("%.3f", fit$beta)`** with a clustered standard error of
`r sprintf("%.3f", fit$se_beta)` — indistinguishable from 1. That is the
reassuring part: the apportionment scales correctly. A city whose predicted base
is twice another's really does have about twice the base. The intercept is not
distinguishable from zero either.

::: {.callout-important}
## The residual spread exceeds the threshold, and the band is widened accordingly

The plan set a tripwire: if the residual SD in logs exceeds about 0.25, say so
and widen the band rather than suppress it. Across all seventeen cities it is
**`r sprintf("%.3f", fit$sigma)`**. Two of the seventeen — Salamanca, whose
population is entirely on Seneca Nation territory, and Ogdensburg, whose
jurisdiction code was created in 2022 and whose implied share is still rising —
have verified structural reasons that do not apply to Albany. Excluding those
two, and only those two, brings it to
**`r sprintf("%.3f", fit_pref$sigma)`** with a slope of
`r sprintf("%.3f", fit_pref$beta)` (se `r sprintf("%.3f", fit_pref$se_beta)`).
That is the calibration used; the all-seventeen figure is carried alongside it
throughout. The memo's §5A works through the reasons city by city.

In plain terms: on the cities where the answer is known, a one-σ miss is
`r sprintf("%.0f", 100*(exp(fit_pref$sigma)-1))`% up or
`r sprintf("%.0f", 100*(1-exp(-fit_pref$sigma)))`% down on the preferred fit, and
`r sprintf("%.0f", 100*(exp(fit$sigma)-1))`% up or
`r sprintf("%.0f", 100*(1-exp(-fit$sigma)))`% down across all seventeen. The band
is multiplicative, so up and down are not equal. The 90% band spans a factor of
`r sprintf("%.1f", exp(2*1.645*fit_pref$sigma))` on the preferred fit. Any
narrower presentation of this estimate would be false precision.
:::

**Salamanca is the largest single residual, and there is a verifiable reason.**
The method over-predicts its base by a factor of more than three. Salamanca sits
on the Allegany Territory of the Seneca Nation: the 2020 Census Block Assignment
Files put **100% of the city's population inside AIANNH area 0080**, against 0%
for Albany and 0% for Olean, the other Cattaraugus city. Retail there is
substantially tribal and outside the state and local tax base, while the
Economic Census counts its receipts. Dropping it lifts $R^2$ to
`r sprintf("%.3f", f_ns$r2)` and cuts $\sigma$ to `r sprintf("%.3f", f_ns$sigma)`.

It is nevertheless **kept in the headline fit**. Excluding an observation after
seeing that it is inconvenient is how error bands get made to look better than
the method deserves. The variant is reported so the reader can see exactly what
that choice costs.

**Dropping Westchester** — the plan asked for this because Westchester's cities
are among Albany's closest structural analogues — moves the slope to
`r sprintf("%.3f", fit_nw$beta)` and *widens* $\sigma$ to
`r sprintf("%.3f", fit_nw$sigma)` on `r fit_nw$n` cities. It does not rescue the
precision.

## 6. The estimate

```{r}
#| label: tbl-est
#| tbl-cap: "From the county base to annual revenue."
tibble(
  Step = c("County taxable base, B_k",
           "Raw apportionment, B_c predicted",
           "Calibrated city base, B_c",
           "× 0.005",
           "× (1 − φ)",
           "× β",
           "Annual revenue"),
  Value = c(sprintf("$%s B", B(B_k)),
            sprintf("$%s B", B(B_pred)),
            sprintf("$%s B", B(est$B_c_central)),
            sprintf("$%s M", M(0.005 * est$B_c_central)),
            sprintf("$%s M", M(0.005 * est$B_c_central * (1 - phi))),
            sprintf("$%s M", M(REV)),
            sprintf("$%s M", M(REV)))) |>
  gt() |> tab_options(table.font.size = pct(92), column_labels.hidden = TRUE)
```

```{r}
#| label: tbl-band
#| tbl-cap: "Central estimate and bands, from the calibration residual spread."
tibble(Interval = c("Central", "68% (±1σ)", "90% (±1.645σ)"),
       `City base ($B)` = c(est$B_c_central, NA, NA),
       `Low ($M)`  = c(NA, rev_of(est$lo68), rev_of(est$lo90)),
       `Revenue ($M)` = c(REV, NA, NA),
       `High ($M)` = c(NA, rev_of(est$hi68), rev_of(est$hi90))) |>
  gt() |>
  fmt_number(`City base ($B)`, decimals = 2) |>
  fmt_number(c(`Low ($M)`, `Revenue ($M)`, `High ($M)`), decimals = 1) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = pct(92))
```

### Consistency with Route B (Validation 7)

The calibrated base implies a city share of `r PC(est$B_c_central/B_k)` and an
$R$ of **`r sprintf("%.2f", R_alb)`**. The seventeen preempting cities range from
`r sprintf("%.2f", min(route_b$R))` to `r sprintf("%.2f", max(route_b$R))` with a
median of `r sprintf("%.2f", median(route_b$R))`. Albany's estimate sits just
below that median, in the company of
`r paste(route_b$city[order(abs(route_b$R - R_alb))][1:3], collapse=", ")`.

That is where it should be. Albany is a city whose county's retail has moved to
the Wolf Road corridor and Crossgates — its retail $R$ is only
`r sprintf("%.2f", 0.2594/POP_SHARE)` — but which holds
`r PC(work_share_ex_pubadmin("3601000","36001"))` of the county's private-sector
jobs. Weak retail pulling down, concentrated employment pulling up, landing
below parity, is the expected shape.

### The preemption comparator

The memo's point that preemption loses is documented here with the same numbers.
Albany County distributes **40% of all 4.00%** to cities and towns on decennial
census population (OSC 2020, Appendix C — a figure the plan carried as an
assumption and which is now verified).

```{r}
#| label: tbl-preempt
#| tbl-cap: "Preempting 1.5 points against the status quo distribution, three-year means."
tibble(Option = c("Status quo: 0.40 × population share × county collections",
                  "Preempt 1.5 points: 0.015 × B_c",
                  "Add on 0.5 points (this estimate): 0.005 × B_c × (1−φ)"),
       `$M` = c(0.40 * POP_SHARE * C_k / 1e6,
                0.015 * est$B_c_central / 1e6,
                REV / 1e6)) |>
  gt() |> fmt_number(`$M`, decimals = 1) |> tab_options(table.font.size = pct(90))
```

Preemption at 1.5 points yields about
\$`r M(0.015*est$B_c_central)` M against \$`r M(0.40*POP_SHARE*C_k)` M from the
existing distribution — it **loses roughly
\$`r M(0.40*POP_SHARE*C_k - 0.015*est$B_c_central)` million a year**, and it
does so for a structural reason: the break-even is
$R = 0.40 \times 4 / 1.5 = `r sprintf("%.4f", albany_breakeven())`$ and Albany's
estimated $R$ is `r sprintf("%.2f", R_alb)`. A city has to be a genuine retail
concentrator for preemption to pay, and Albany is not one. The add-on, which
leaves the distribution untouched, does not face this trade-off at all.

## 7. Sensitivity (Validation 6)

```{r}
#| label: tbl-sens
#| tbl-cap: "Each row changes one modelling choice and re-runs the whole apportionment and calibration. Revenue is the calibrated central estimate."
variants <- tribble(
  ~Variant, ~args,
  "Central", list(),
  "Business allocator: total employment incl. public administration", list(business_allocator = "lodes"),
  "Business allocator: Econ Census payroll",         list(business_allocator = "ec_payroll"),
  "Delivered goods 0% to residence",                 list(delivered_to_residence = 0),
  "Delivered goods 100% to residence",               list(delivered_to_residence = 1),
  "Residence allocator: households not income",      list(resid_var = "hh"),
  "E-commerce: 15% of prone groups to residence",    list(ecommerce_frac = 0.15),
  "E-commerce: 30% of prone groups to residence",    list(ecommerce_frac = 0.30),
  "NAICS 4413 sourced to residence",                 list(mv_include_4413 = TRUE),
  "Motor vehicles: ACS vehicles available",          list(mv_allocator = "acs_vehicles"))

sens <- variants |>
  rowwise() |>
  mutate(B_pred_v = do.call(apportion_variant, c(ALB, args)),
         B_cal_v  = apply_calibration(fit_pref, B_pred_v)$B_c_central,
         Rev      = rev_of(B_cal_v)) |>
  ungroup()

sens |>
  transmute(Variant,
            `Predicted base ($B)` = B_pred_v / 1e9,
            `Calibrated base ($B)` = B_cal_v / 1e9,
            `Revenue ($M)` = Rev / 1e6,
            `vs central` = Rev / REV - 1) |>
  gt() |>
  fmt_number(c(`Predicted base ($B)`, `Calibrated base ($B)`), decimals = 3) |>
  fmt_number(`Revenue ($M)`, decimals = 1) |>
  fmt_percent(`vs central`, decimals = 1) |>
  tab_options(table.font.size = pct(85))
```

The striking result is how little most of these matter. The delivered-goods
split — the plan's flagged 50/50 assumption — moves the answer by less than a
tenth of a percent, because the Economic Census store share and the ACS income
share for those groups happen to be nearly identical. The e-commerce assignment,
which the loss of NAICS 454 made unresolvable, moves it by well under 1% for the
same reason: reassigning receipts from a `r PC(0.2594)` allocator to a
`r PC(resid_share("01000","001","agginc"))` allocator barely changes anything.
**The unresolvable data problem turned out to be immaterial.**

What does matter is the **business allocator**, and only one option there is
defensible. Counting government jobs raises the estimate by
`r sprintf("%.0f", 100*(sens$Rev[2]/REV-1))`%, but government purchases are
exempt (Tax Law §1116(a)(1)), so using government employment to apportion
taxable business purchases means distributing a taxable base by a measure of
exempt activity. The row is reported because it was the project plan's original
choice, not as a credible alternative. The calibration cities have too little
government employment to test it either way. Using the school district's
directly measured share for the utilities class instead of the allocator changes
the estimate by well under 1%.

### Behavioural response, β

The add-on creates an 8.5% rate inside the city against 8.0% outside — a
6.25% increase in the tax, or 0.46% on a tax-inclusive price. Motor vehicles are
unaffected, since they are sourced to the purchaser's residence regardless of
where the dealer sits. Business purchases, utilities and services are largely
immobile. Only the store classes, about
`r PC(sum(appo$B_kg[appo$sourcing_class %in% c("store","delivered_split")])/B_k)`
of the base, are plausibly mobile at all, and only near the city line.

$\beta = 1.0$ is carried as central. A low case of
$\beta = `r sprintf("%.2f", BETA_LO)`$ — a 3% loss of base against a 6.25%
increase in the tax, i.e. an elasticity of about 0.5 with respect to the *tax
rate*; with respect to the tax-inclusive price, which rises only 0.46%, it is
about 6.5, which is why this is a deliberately aggressive low case — gives
\$`r M(rev_of(est$B_c_central, BETA_LO))` M, a
`r sprintf("%.1f", 100*(1-BETA_LO))`% reduction. This is small relative to the
calibration band and does not change any conclusion.

The cross-border literature establishes that such responses exist and are
concentrated within a few miles of a rate discontinuity (Agrawal, 2015; Baker,
Johnson & Kueng, 2021). Neither paper studies a 0.5-point intra-county
differential in an upstate New York city, so **the specific low-case elasticity
above is illustrative** rather than transported from a published estimate, and
is labelled as such.

## 8. Timing and first-year cash

Local rate changes take effect only on 1 March, 1 June, 1 September or
1 December, and require 90 days' notice to the Commissioner (Tax Law §1210(d),
§1211). Vendors get transitional rules for straddling contracts and
subscriptions (Tax Bulletin ST-895). Distributions then lag liability by one to
two months, and the first two months of each quarter are paid as estimates and
trued up in the third.

Assuming a city fiscal year ending 31 December and an average cash lag of
1.5 months:

```{r}
#| label: tbl-timing
#| tbl-cap: "First-fiscal-year cash under the two most likely start dates, at the central annual estimate."
tibble(`Start date` = c("1 March", "1 December"),
       `Months of liability in first FY` = c(10, 1),
       `Months of cash in first FY` = c(8.5, 0),
       `First-year cash ($M)` = c(REV * 8.5/12, 0),
       `Steady state ($M)` = c(REV, REV) / 1e6) |>
  mutate(`First-year cash ($M)` = `First-year cash ($M)` / 1e6) |>
  gt() |>
  fmt_number(c(`First-year cash ($M)`, `Steady state ($M)`), decimals = 1) |>
  tab_options(table.font.size = pct(90))
```

A 1 March start captures roughly \$`r M(REV*8.5/12)` M of cash in the first
calendar year and reaches the full \$`r M(REV)` M in the second. A 1 December
start yields essentially nothing in the first calendar year — the single month of
December liability arrives as cash in mid-January. If the objective is revenue
in a particular budget year, the start date is worth as much as several of the
modelling choices in §7.

## 9. Validation summary

| # | Validation | Result |
|---|---|---|
| 1 | 4% × county base vs county distributions | Gap `r PC(min(-phi_tab$phi),1)` to `r PC(max(-phi_tab$phi),1)`; passes (plan threshold 10%) |
| 2 | Economic Census places vs county total | Places cover 99.2% of county retail receipts; residual 0.8% |
| 3 | Utilities: allocator vs measured school-district base | allocator `r PC(work_share_ex_pubadmin("3601000","36001"))`, measured `r PC(albany_measured_utility_share())`, total employment `r PC(work_share("3601000","36001"))` — measured lies between |
| 4 | Motor vehicles: DMV vs ACS income share | `r PC(mv_share("3601000","ALBANY"))` vs `r PC(resid_share("01000","001","agginc"))`; DMV lower, consistent with lower urban vehicle ownership |
| 5 | Calibration fit | Preferred (15 cities): β = `r sprintf("%.3f", fit_pref$beta)` (se `r sprintf("%.3f", fit_pref$se_beta)`), σ = `r sprintf("%.3f", fit_pref$sigma)`. All 17: σ = `r sprintf("%.3f", fit$sigma)`, **above the 0.25 threshold the plan set for the full sample** |
| 6 | Sensitivity table | §7; business allocator dominates, everything else < 1% |
| 7 | Albany's R against the Route B distribution | `r sprintf("%.2f", R_alb)` against a range of `r sprintf("%.2f–%.2f", min(route_b$R), max(route_b$R))`, median `r sprintf("%.2f", median(route_b$R))` — inside |

## 10. What would move this estimate

Honest ranking of what the reader should worry about.

1. **The exempt public sector.** `r PC(pubadmin_share("3601000","36001"))` of
   county public-administration employment is in the city. Government purchases
   are exempt, so those jobs are excluded from the business allocator; including
   them would raise revenue by
   `r sprintf("%.0f", 100*(sens$Rev[2]/REV-1))`%, and the calibration cannot
   adjudicate. Hospitals and universities, largely exempt too, remain in the
   count. This is the largest real uncertainty.
2. **The calibration spread itself.** σ = `r sprintf("%.3f", fit_pref$sigma)` on
   the preferred fit means a one-σ miss of
   `r sprintf("%.0f", 100*(exp(fit_pref$sigma)-1))`% up or
   `r sprintf("%.0f", 100*(1-exp(-fit_pref$sigma)))`% down on cities where the
   truth is known (`r sprintf("%.0f", 100*(exp(fit$sigma)-1))`% /
   `r sprintf("%.0f", 100*(1-exp(-fit$sigma)))`% across all seventeen). Albany may
   be an easier case than Salamanca or Ogdensburg — it is large, and its
   allocators resolve at fine NAICS detail — but nothing in the data proves that.
3. **Economic Census vintage and basis.** EC 2022 is a single-year snapshot
   applied to `r min(ANALYSIS_FYS)-1`–`r max(ANALYSIS_FYS)` bases. Because the
   allocators are shares, level drift matters much less than composition drift.
   More importantly, EC receipts are *gross*, not taxable, and are recorded at
   the establishment's location; the assumption is that the taxable fraction of a
   NAICS group is similar inside and outside the city.
4. **Remote sales.** DTF does not publish taxable sales by delivery
   jurisdiction. The residence allocator for delivered goods is an assumption,
   though §7 shows the estimate barely responds to it.
5. **Ogdensburg's rate**, one of seventeen calibration observations, rests on
   footnote reasoning rather than a published rate table (see Phase 2). It is
   small and does not drive the fit.

## Collections divided by rate against reported taxable sales, by county

```{r}
#| label: tbl-county-test-full
#| tbl-cap: "For each county outside New York City: reported taxable sales, collections (plus amounts preempted by its cities) divided by the county rate, and the ratio; three-year means and the range across FY2023–2025. County rates from Publication 718."
R6 <- readRDS(here::here("data", "processed", "results.rds"))
R6$county_test_by_county |>
  transmute(County = county, `County rate %` = r_k_pub718, `Taxing cities` = n_cities,
            `Reported base, $ million` = B_k / 1e6, `Collections ÷ rate, $ million` = inferred / 1e6,
            `Ratio` = ratio, `Lowest year` = ratio_min, `Highest year` = ratio_max) |>
  knitr::kable(format = "pipe", digits = c(0, 3, 0, 0, 0, 3, 3, 3), align = "lrrrrrrr", format.args = list(big.mark = ","))
```

## Residential energy: county-by-county test

Whether the DTF county row contains residential energy, tested by grouping
counties by their Publication 718-R status (see the apportionment chapter).

```{r}
#| label: tbl-resid-counties
#| tbl-cap: "Taxable sales per person by county, three-year mean, with residential-energy tax status from Publication 718-R."
R6 <- readRDS(here::here("data", "processed", "results.rds"))
R6$re_test |> arrange(status, desc(util_pc)) |>
  transmute(County = county, `Residential energy tax` = status, Population = P,
            `Electric 2211, $/person` = elec_pc, `Utilities 22, $/person` = util_pc,
            `Petroleum 4247, $/person` = petrol_pc, `Whole base, $/person` = total_pc) |>
  knitr::kable(format = "pipe", digits = 0, align = "llrrrrr", format.args = list(big.mark = ","))
```

## References

Agrawal, D. R. (2015). The tax gradient: Spatial aspects of fiscal competition.
*American Economic Journal: Economic Policy, 7*(2), 1–29.
<https://www.aeaweb.org/articles?id=10.1257/pol.20120360>

Baker, S. R., Johnson, S., & Kueng, L. (2021). Shopping for lower sales tax
rates. *American Economic Journal: Macroeconomics, 13*(3), 209–250.
<https://www.aeaweb.org/articles?id=10.1257/mac.20190026>

New York State Department of Taxation and Finance. (2026). *Taxable sales and purchases quarterly data: Beginning sales tax year 2013–2014* [Data set]. <https://data.ny.gov/Government-Finance/Taxable-Sales-And-Purchases-Quarterly-Data-Beginni/ny73-2j3u/about_data>

New York State Department of Taxation and Finance. (2025). *State and local sales tax distributions: Beginning fiscal years ended March 31, 1995* [Data set]. <https://data.ny.gov/Government-Finance/State-and-Local-Sales-Tax-Distributions-Beginning-/5g2s-tnb7>

New York State Department of Taxation and Finance. (2026). *AS001—Revenue distribution certification*. <https://www.tax.ny.gov/research/stats/statistics/sales_tax/government/as001.htm>

New York State Department of Taxation and Finance. (2025). *Publication 718: New York State sales and use tax rates by jurisdiction*. <https://www.tax.ny.gov/pdf/publications/sales/pub718.pdf>

New York State Department of Taxation and Finance. (2025). *Publication 718-A: Enactment and effective dates of sales and use tax rates*. <https://www.tax.ny.gov/pdf/publications/sales/pub718a.pdf>

New York State Department of Taxation and Finance. (2025). *Publication 718-C: Sales and use tax rates on clothing and footwear*. <https://www.tax.ny.gov/pdf/publications/sales/pub718c.pdf>

New York State Department of Taxation and Finance. (1990). *TSB-M-90(6)S: The Albany City School District imposes a 3% sales and use tax on utilities and utility services*. <https://tax.ny.gov/pdf/memos/sales/m90_6s.pdf>

New York State Department of Taxation and Finance. (2014). *Tax Bulletin ST-895: Transitional provisions for sales tax rate changes*. <https://www.tax.ny.gov/pubs_and_bulls/tg_bulletins/st/transitional_provisions.htm>

New York State Department of Motor Vehicles. (2026). *Vehicle, snowmobile, and boat registrations* [Data set]. Open NY. <https://data.ny.gov/Transportation/Vehicle-Snowmobile-and-Boat-Registrations/w4pv-hbkt>

N.Y. Tax Law §§ 1210, 1211, 1214, 1223, 1224.

Office of the New York State Comptroller. (2020). *Understanding local government sales tax in New York State: 2020 update*. <https://www.osc.ny.gov/files/local-government/publications/pdf/understanding-local-government-sales-tax-in-nys-2020-update.pdf>

U.S. Census Bureau. (2021). *2020 Census redistricting data (P.L. 94-171) summary file* [API dataset `dec/pl`]. <https://api.census.gov/data/2020/dec/pl.html>

U.S. Census Bureau. (2021). *2020 Census block assignment files*. <https://www2.census.gov/geo/docs/maps-data/data/baf2020/>

U.S. Census Bureau. (2024). *American Community Survey 5-year estimates: 2023* [API dataset `acs/acs5`]. <https://api.census.gov/data/2023/acs/acs5.html>

U.S. Census Bureau. (2025). *County Business Patterns: 2023* [API dataset `cbp`]. <https://api.census.gov/data/2023/cbp.html>

U.S. Census Bureau. (2025). *Economic Census: Summary statistics for the U.S., states, and selected geographies: 2022* [API dataset `ecnbasic`]. <https://api.census.gov/data/2022/ecnbasic.html>

U.S. Census Bureau. (n.d.). *2020 relationship files: ZCTA to place, ZCTA to block*. <https://www.census.gov/geographies/reference-files/time-series/geo/relationship-files.html>

U.S. Census Bureau, Longitudinal Employer-Household Dynamics. (2025). *LEHD Origin-Destination Employment Statistics (LODES) version 8: Workplace Area Characteristics, New York, 2023*. <https://lehd.ces.census.gov/data/>

---

*Every figure in this document is computed at render time from cached pulls in
`data/raw/` by `R/00_setup.R` … `R/05_calibrate.R`. Provenance for each raw file
— source URL, pull timestamp, SHA-256 — is in `data/raw/MANIFEST.csv`. The only
hand-entered inputs are the two crosswalk tables in `data/crosswalk/`, each
carrying a source URL per row. See `docs/data-layouts.md` for the verified schema
of every dataset and the ten places where it differed from what was assumed.*
