---
title: "Revenue from a 0.5% City of Albany add-on sales tax"
subtitle: "Apportionment of the county base, calibrated against New York's preempting cities"
date: today
---

```{r setup}
#| include: false
library(here)
source(here("R", "04_allocate.R"))
suppressPackageStartupMessages({library(gt); library(ggplot2)})

ALB <- list("Albany city, New York", "3601000", "36001", "ALBANY", "ALBANY")
POP_SHARE <- {
  p <- read_pop()
  p$place$P_c[p$place$place_fips == "3601000"] / p$county$P_k[p$county$county_fips == "36001"]
}

appo   <- do.call(apportion_city, ALB)
B_k    <- sum(appo$B_kg)
B_pred <- sum(appo$B_cg, na.rm = TRUE)

calib  <- build_calibration()
fit    <- fit_calibration(calib)
fit_nw <- fit_calibration(calib, drop_westchester = TRUE)
est    <- apply_calibration(fit, B_pred)

phi_tab <- albany_phi()
phi     <- mean(phi_tab$phi)
BETA    <- 1.0
BETA_LO <- 0.97

rev_of <- function(b, beta = BETA) revenue_from_base(b, phi = phi, beta = beta)
REV     <- rev_of(est$B_c_central)

route_b <- summarise_route_b(build_route_b())
R_alb   <- (est$B_c_central / B_k) / POP_SHARE
R_raw   <- (B_pred / B_k) / POP_SHARE

C_k <- mean(read_distributions() |>
              filter(taxing_jurisdiction == "Albany County Sales and Use Tax",
                     fy %in% ANALYSIS_FYS) |> pull(amt))

M  <- function(x, d = 1) sprintf(paste0("%.", d, "f"), x / 1e6)
B  <- function(x, d = 2) sprintf(paste0("%.", d, "f"), x / 1e9)
PC <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
```

::: {.callout-note}
## The estimate

A 0.5 percentage-point City of Albany add-on sales and use tax, on the same base
Albany County taxes at 4%, destination-sourced, layered on the existing 8% with
no preemption and no change to county distributions, would raise approximately

### **\$`r M(REV)` million a year** {.unnumbered}

with a 68% range of **\$`r M(est$lo68 * 0.005 * (1-phi))`–`r M(est$hi68 * 0.005 * (1-phi))` million**
and a 90% range of **\$`r M(est$lo90 * 0.005 * (1-phi))`–`r M(est$hi90 * 0.005 * (1-phi))` million**,
in the price and activity levels of sales tax years `r min(ANALYSIS_FYS)-1`-`r max(ANALYSIS_FYS)`.

The band is wide, and honestly so: it is driven by how poorly the method
performs on the cities where the answer can be checked, not by sampling noise.
:::

## The identity

$$\text{Revenue} = 0.005 \times B_c \times (1 - \varphi) \times \beta$$

| term | meaning | value used |
|---|---|---|
| $B_c$ | City of Albany taxable sales and purchases | **\$`r B(est$B_c_central)` B** (calibrated) |
| $\varphi$ | administrative deduction, $1 - \frac{\text{county distributions}}{0.04 \times \text{county base}}$ | **`r PC(phi, 2)`** |
| $\beta$ | behavioral response to the 0.5-point differential | **`r sprintf("%.2f", BETA)`** central |

$B_c$ is published by nobody. Everything below is about estimating it and being
honest about how well that can be done.

## 1. The county base, and the reconciliation

Albany County's taxable sales and purchases come from the DTF quarterly file,
summed over all `r nrow(appo)` NAICS groups and averaged over the three sales
tax years ending `r max(ANALYSIS_FYS)`: **\$`r B(B_k)` billion**.

Validation 1 in §9 requires that 4% of that base reconcile to what the county
actually received.

```{r}
#| label: tbl-recon
#| tbl-cap: "Validation 1. County distributions against 4% of the published county base. The sales tax year runs March–February; the state fiscal year runs April–March, so the two are offset by one month."
phi_tab |>
  transmute(`Fiscal year` = fy,
            `County base ($B)` = B / 1e9,
            `4% of base ($M)` = 0.04 * B / 1e6,
            `Distributions ($M)` = C / 1e6,
            `Gap` = -phi) |>
  gt() |>
  fmt_number(`County base ($B)`, decimals = 3) |>
  fmt_number(c(`4% of base ($M)`, `Distributions ($M)`), decimals = 1) |>
  fmt_percent(Gap, decimals = 2) |>
  tab_options(table.font.size = pct(90))
```

The gap is under 2.5% in every year and averages `r PC(-phi, 2)` — the plan
called for "low single digits percent" and flagged anything over 10% as a
broken pull. It passes.

The gap is also *negative on average*: the county received slightly more than
4% of the published base. That is not an administrative surplus. It reflects the
one-month calendar offset, late and amended returns, use tax on purchases that
does not sit in the "taxable sales" measure, and the withheld cells discussed
below. The practical consequence is that **$\varphi$ is indistinguishable from
zero**, and it is carried at its observed mean of `r PC(phi, 2)` rather than
being invented.

::: {.callout-warning}
## Cell suppression

DTF withholds individual county × NAICS cells for disclosure, and `ny73-2j3u`
has no jurisdiction total row. Every county base here is therefore the sum of
its *published* groups. For Albany the withheld share is about 3.6% of groups;
the reconciliation above bounds their aggregate effect, since collections
reconcile to the published base within a couple of percent.
:::

## 2. Sourcing classes

Every one of the `r nrow(appo)` NAICS groups maps to exactly one class, recorded
with a rationale and a source in `data/crosswalk/naics_sourcing_class.csv`.

```{r}
#| label: tbl-classes
#| tbl-cap: "Sourcing classes and the allocator each uses. Base is the three-year mean county base."
appo |>
  group_by(sourcing_class) |>
  summarise(`Groups` = n(),
            `County base ($M)` = sum(B_kg) / 1e6,
            `Share of base` = sum(B_kg) / B_k,
            `Effective a_g` = sum(B_cg, na.rm = TRUE) / sum(B_kg),
            `City base ($M)` = sum(B_cg, na.rm = TRUE) / 1e6, .groups = "drop") |>
  mutate(Allocator = c(business = "LODES workplace employment",
                       delivered_split = "50% Econ Census receipts / 50% ACS income",
                       motor_vehicle = "DMV resident registrations",
                       store = "Econ Census receipts",
                       utilities = "LODES workplace employment")[sourcing_class],
         Class = sourcing_class) |>
  select(Class, Allocator, Groups, `County base ($M)`, `Share of base`,
         `Effective a_g`, `City base ($M)`) |>
  arrange(desc(`County base ($M)`)) |>
  gt() |>
  fmt_number(c(`County base ($M)`, `City base ($M)`), decimals = 1) |>
  fmt_percent(`Share of base`, decimals = 1) |>
  fmt_number(`Effective a_g`, decimals = 3) |>
  tab_options(table.font.size = pct(85))
```

Two choices deserve comment.

**Motor vehicles are 4411 and 4412 only.** Tax Law §1214 sources sales of
vehicles required to be registered to the purchaser's residence, so the dealer's
location carries no information about where the tax lands. That reasoning does
not extend to 4413 (parts, accessories and tires), which is ordinary
over-the-counter retail and is left store-based. The project plan specified all
of NAICS 441; 4413 is \$65 M of a \$`r B(B_k)` B base and moving it changes the
answer by less than a tenth of a percent — it is in the sensitivity table.

**There is no e-commerce class**, because NAICS 2022 eliminated sector 454 and
folded electronic shopping into the store categories. The plan assumed a
nonstore group to allocate by residence; no such group exists in the analysis
window and DTF publishes no split. This turns out not to matter much, and §7
shows why.

## 3. The allocators

```{r}
#| label: tbl-alloc
#| tbl-cap: "City share of the county under each allocator, with the population share for reference."
tibble(
  Allocator = c("LODES workplace employment (business, utilities)",
                "  — of which public administration",
                "  — excluding public administration",
                "Economic Census payroll, matched sectors",
                "Economic Census receipts, store classes (base-weighted)",
                "DMV resident registrations, model year ≥ 2024",
                "ACS aggregate household income (residence)",
                "ACS households",
                "Measured: school district utility tax ÷ 0.03",
                "Population (2020 Census)"),
  Share = c(work_share("3601000", "36001"),
            pubadmin_share("3601000", "36001"),
            work_share_ex_pubadmin("3601000", "36001"),
            ec_payroll_share("Albany city, New York", "001"),
            with(appo[appo$sourcing_class %in% c("store", "delivered_split"), ],
                 sum(a_store_used * B_kg) / sum(B_kg)),
            mv_share("3601000", "ALBANY"),
            resid_share("01000", "001", "agginc"),
            resid_share("01000", "001", "hh"),
            albany_measured_utility_share(),
            POP_SHARE)) |>
  gt() |>
  fmt_percent(Share, decimals = 1) |>
  tab_options(table.font.size = pct(88))
```

The spread across allocators is the whole story of this estimate. Albany city
holds **`r PC(work_share("3601000","36001"))` of the county's jobs** but only
**`r PC(mv_share("3601000","ALBANY"))` of its vehicle registrations** — a
factor of two and a half, around a population share of `r PC(POP_SHARE)`. Which
allocator applies to which part of the base determines the answer.

Two results are worth pulling out.

**State government.** `r PC(pubadmin_share("3601000","36001"))` of Albany
County's public-administration jobs are inside the city, and public
administration is
`r PC(with(lodes_place |> filter(cty=="36001"), sum(CNS20)/sum(C000)))` of all
county employment. The Capitol complex is the single largest concentrated
taxable purchaser in the county and it is invisible to the Economic Census. This
is why LODES rather than Economic Census payroll is the central business
allocator, and it is the biggest single sensitivity in §7.

**Validation 3 passes.** The utilities class is allocated in the generic
apportionment by workplace employment, giving
`r PC(work_share("3601000","36001"))`. It can also be *measured*: the Albany
City School District's 3% utility tax implies a city utility base of
`r PC(albany_measured_utility_share())` of the county's NAICS 22 + 517 base. The
allocator and the measurement agree to within
`r sprintf("%.0f", abs(100*(work_share("3601000","36001")/albany_measured_utility_share()-1)))`%.
That is the only place in this exercise where an allocator can be checked
against a direct observation of an Albany city base, and it holds.

**Validation 4.** The DMV registration share
(`r PC(mv_share("3601000","ALBANY"))`) sits below the ACS income share
(`r PC(resid_share("01000","001","agginc"))`) and well below population. Albany
city households own fewer vehicles per person than the county average — an urban
core with students, transit and lower incomes. Registrations carrying
out-of-state ZIPs (Ft Lauderdale, Port Newark: leasing and fleet registrants,
`r PC(with(dmv_by_zip |> filter(county=="ALBANY"), sum(n[state!="NY"])/sum(n)))`
of the county total) have no in-county residence to assign and are dropped from
both numerator and denominator, which distributes them in proportion to the
resident pattern.

::: {.callout-note}
## ZIP codes were not assigned whole

Albany's ZIPs straddle Colonie, Guilderland and Menands. The ZCTA-to-place
relationship file publishes land area only, and area weighting badly understates
a dense central city: ZCTA 12203 is 40% of Albany city by area but
**62% by population**. Exact population weights were built instead from 2020
Census block populations joined to block→place (Block Assignment Files) and
block→ZCTA (the national ZCTA-to-block relationship file, stream-filtered to the
13 counties in this analysis rather than downloaded whole).
:::

## 4. The raw apportionment

$$B_c^{\text{pred}} = \sum_g B_{k,g} \times a_g = \textbf{\$`r B(B_pred)` billion}$$

which is `r PC(B_pred/B_k)` of the county base against a population share of
`r PC(POP_SHARE)` — an implied $R$ of `r sprintf("%.2f", R_raw)`.

```{r}
#| label: tbl-groups
#| tbl-cap: "The twenty largest NAICS groups by county base, showing the allocator resolved for each and the fallback flag."
appo |>
  arrange(desc(B_kg)) |>
  head(20) |>
  transmute(NAICS = naics_industry_group,
            Description = substr(description, 1, 44),
            Class = sourcing_class,
            `County base ($M)` = B_kg / 1e6,
            Allocator = allocator,
            `a_g` = a_g,
            `City base ($M)` = B_cg / 1e6,
            Fallback = ifelse(fallback, "yes", "")) |>
  gt() |>
  fmt_number(c(`County base ($M)`, `City base ($M)`), decimals = 1) |>
  fmt_number(`a_g`, decimals = 3) |>
  tab_options(table.font.size = pct(78))
```

Of the `r sum(appo$sourcing_class %in% c("store","delivered_split"))` groups
needing an Economic Census share,
`r sum(grepl("4-digit", appo$store_method))` resolved at the 4-digit level and
`r sum(grepl("3-digit", appo$store_method))` fell back to 3-digit. None needed
the sector level and none went unresolved.

## 5. Calibration

The apportionment above is a construction, not a measurement. The seventeen
cities from Phase 2 are the places where the same construction can be run and
then checked against a base that is actually observed.

```{r}
#| label: tbl-calib
#| tbl-cap: "The identical apportionment run on every calibration city, against the base implied by its own collections. Ratio below 1 means the method over-predicts."
calib |>
  transmute(City = city, County = county,
            `Observed ($M)` = B_c_obs / 1e6,
            `Predicted ($M)` = B_c_pred / 1e6,
            `Obs ÷ Pred` = ratio) |>
  arrange(`Obs ÷ Pred`) |>
  gt() |>
  fmt_number(c(`Observed ($M)`, `Predicted ($M)`), decimals = 0) |>
  fmt_number(`Obs ÷ Pred`, decimals = 3) |>
  data_color(columns = `Obs ÷ Pred`, palette = c("#d6604d", "#f7f7f7", "#4393c3"),
             domain = c(0.25, 1.25)) |>
  tab_options(table.font.size = pct(88))
```

Fitting $\log B_c^{\text{obs}} = \alpha + \beta \log B_c^{\text{pred}}$, with
standard errors clustered by county because Fulton, Cattaraugus, Oneida and
Westchester each contribute more than one city:

```{r}
#| label: tbl-fit
#| tbl-cap: "Calibration fits. Sigma is the residual standard deviation in logs and sets the width of the band."
mk <- function(lbl, f) tibble(Specification = lbl, n = f$n, Clusters = f$n_clusters,
                              `α` = f$alpha, `se(α)` = f$se_alpha,
                              `β` = f$beta, `se(β)` = f$se_beta,
                              `σ (logs)` = f$sigma, `R²` = f$r2)
c_ns  <- calib |> filter(city != "Salamanca")
f_ns  <- { d <- c_ns; m <- lm(log_obs ~ log_pred, d)
           ct <- lmtest::coeftest(m, vcov. = sandwich::vcovCL(m, cluster = d$county))
           list(n = nrow(d), n_clusters = n_distinct(d$county), alpha = coef(m)[1],
                se_alpha = ct[1,2], beta = coef(m)[2], se_beta = ct[2,2],
                sigma = summary(m)$sigma, r2 = summary(m)$r.squared) }
bind_rows(mk("All 17 cities (used)", fit),
          mk("Excluding Westchester", fit_nw),
          mk("Excluding Salamanca", f_ns)) |>
  gt() |>
  fmt_number(c(`α`, `se(α)`, `β`, `se(β)`), decimals = 3) |>
  fmt_number(c(`σ (logs)`, `R²`), decimals = 3) |>
  tab_options(table.font.size = pct(85))
```

```{r}
#| label: fig-calib
#| fig-cap: "Observed against predicted city base, log scale. The dashed line is the fitted calibration, the grey line is equality. Albany's prediction is marked on the horizontal axis and its calibrated estimate on the vertical."
#| fig-height: 5.5
ggplot(calib, aes(B_c_pred / 1e6, B_c_obs / 1e6)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey75") +
  geom_abline(slope = fit$beta, intercept = fit$alpha - (fit$beta - 1) * log(1e6),
              linetype = "dashed", colour = "#b2182b") +
  geom_point(aes(colour = westchester), size = 2.8) +
  ggrepel::geom_text_repel(aes(label = city), size = 2.9, max.overlaps = 20, seed = 1) +
  annotate("point", x = B_pred / 1e6, y = est$B_c_central / 1e6,
           shape = 18, size = 5.5, colour = "#1b7837") +
  annotate("text", x = B_pred / 1e6, y = est$B_c_central / 1e6,
           label = "  Albany (calibrated)", hjust = 0, size = 3.2, colour = "#1b7837") +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c("FALSE" = "#4d4d4d", "TRUE" = "#762a83"),
                      labels = c("Rest of state", "Westchester"), name = NULL) +
  labs(x = "Predicted city base, $M (log scale)",
       y = "Observed city base, $M (log scale)") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

### Reading the fit

The slope is **`r sprintf("%.3f", fit$beta)`** with a clustered standard error of
`r sprintf("%.3f", fit$se_beta)` — indistinguishable from 1. That is the
reassuring part: the apportionment scales correctly. A city whose predicted base
is twice another's really does have about twice the base. The intercept is not
distinguishable from zero either.

::: {.callout-important}
## The residual spread exceeds the threshold, and the band is widened accordingly

The plan set a tripwire: if the residual SD in logs exceeds about 0.25, say so
and widen the band rather than suppress it. It is
**`r sprintf("%.3f", fit$sigma)`**.

In plain terms: on the cities where the answer is known, this method is
routinely wrong by a third in either direction. The 68% band below is
`r sprintf("%.0f", 100*(exp(fit$sigma)-1))`% up and
`r sprintf("%.0f", 100*(1-exp(-fit$sigma)))`% down, and the 90% band is roughly a
factor of `r sprintf("%.1f", exp(1.645*fit$sigma))` from bottom to top. Any
narrower presentation of this estimate would be false precision.
:::

**Salamanca is the largest single residual, and there is a verifiable reason.**
The method over-predicts its base by a factor of more than three. Salamanca sits
on the Allegany Territory of the Seneca Nation: the 2020 Census Block Assignment
Files put **100% of the city's population inside AIANNH area 0080**, against 0%
for Albany and 0% for Olean, the other Cattaraugus city. Retail there is
substantially tribal and outside the state and local tax base, while the
Economic Census counts its receipts. Dropping it lifts $R^2$ to
`r sprintf("%.3f", f_ns$r2)` and cuts $\sigma$ to `r sprintf("%.3f", f_ns$sigma)`.

It is nevertheless **kept in the headline fit**. Excluding an observation after
seeing that it is inconvenient is how error bands get made to look better than
the method deserves. The variant is reported so the reader can see exactly what
that choice costs.

**Dropping Westchester** — the plan asked for this because Westchester's cities
are among Albany's closest structural analogues — moves the slope to
`r sprintf("%.3f", fit_nw$beta)` and *widens* $\sigma$ to
`r sprintf("%.3f", fit_nw$sigma)` on `r fit_nw$n` cities. It does not rescue the
precision.

## 6. The estimate

```{r}
#| label: tbl-est
#| tbl-cap: "From the county base to annual revenue."
tibble(
  Step = c("County taxable base, B_k",
           "Raw apportionment, B_c predicted",
           "Calibrated city base, B_c",
           "× 0.005",
           "× (1 − φ)",
           "× β",
           "Annual revenue"),
  Value = c(sprintf("$%s B", B(B_k)),
            sprintf("$%s B", B(B_pred)),
            sprintf("$%s B", B(est$B_c_central)),
            sprintf("$%s M", M(0.005 * est$B_c_central)),
            sprintf("$%s M", M(0.005 * est$B_c_central * (1 - phi))),
            sprintf("$%s M", M(REV)),
            sprintf("$%s M", M(REV)))) |>
  gt() |> tab_options(table.font.size = pct(92), column_labels.hidden = TRUE)
```

```{r}
#| label: tbl-band
#| tbl-cap: "Central estimate and bands, from the calibration residual spread."
tibble(Interval = c("Central", "68% (±1σ)", "90% (±1.645σ)"),
       `City base ($B)` = c(est$B_c_central, NA, NA),
       `Low ($M)`  = c(NA, rev_of(est$lo68), rev_of(est$lo90)),
       `Revenue ($M)` = c(REV, NA, NA),
       `High ($M)` = c(NA, rev_of(est$hi68), rev_of(est$hi90))) |>
  gt() |>
  fmt_number(`City base ($B)`, decimals = 2) |>
  fmt_number(c(`Low ($M)`, `Revenue ($M)`, `High ($M)`), decimals = 1) |>
  sub_missing(missing_text = "") |>
  tab_options(table.font.size = pct(92))
```

### Consistency with Route B (Validation 7)

The calibrated base implies a city share of `r PC(est$B_c_central/B_k)` and an
$R$ of **`r sprintf("%.2f", R_alb)`**. The seventeen preempting cities range from
`r sprintf("%.2f", min(route_b$R))` to `r sprintf("%.2f", max(route_b$R))` with a
median of `r sprintf("%.2f", median(route_b$R))`. Albany's estimate sits just
below that median, in the company of
`r paste(route_b$city[order(abs(route_b$R - R_alb))][1:3], collapse=", ")`.

That is where it should be. Albany is a city whose county's retail has moved to
the Wolf Road corridor and Crossgates — its retail $R$ is only
`r sprintf("%.2f", 0.2594/POP_SHARE)` — but which contains the State Capitol and
`r PC(work_share("3601000","36001"))` of county employment. Weak retail pulling
down, concentrated employment pulling up, landing near parity, is the expected
shape.

### The preemption comparator

The memo's point that preemption loses is documented here with the same numbers.
Albany County distributes **40% of all 4.00%** to cities and towns on decennial
census population (OSC 2020, Appendix C — a figure the plan carried as an
assumption and which is now verified).

```{r}
#| label: tbl-preempt
#| tbl-cap: "Preempting 1.5 points against the status quo distribution, three-year means."
tibble(Option = c("Status quo: 0.40 × population share × county collections",
                  "Preempt 1.5 points: 0.015 × B_c",
                  "Add on 0.5 points (this estimate): 0.005 × B_c × (1−φ)"),
       `$M` = c(0.40 * POP_SHARE * C_k / 1e6,
                0.015 * est$B_c_central / 1e6,
                REV / 1e6)) |>
  gt() |> fmt_number(`$M`, decimals = 1) |> tab_options(table.font.size = pct(90))
```

Preemption at 1.5 points yields about
\$`r M(0.015*est$B_c_central)` M against \$`r M(0.40*POP_SHARE*C_k)` M from the
existing distribution — it **loses roughly
\$`r M(0.40*POP_SHARE*C_k - 0.015*est$B_c_central)` million a year**, and it
does so for a structural reason: the break-even is
$R = 0.40 \times 4 / 1.5 = `r sprintf("%.4f", albany_breakeven())`$ and Albany's
estimated $R$ is `r sprintf("%.2f", R_alb)`. A city has to be a genuine retail
concentrator for preemption to pay, and Albany is not one. The add-on, which
leaves the distribution untouched, does not face this trade-off at all.

## 7. Sensitivity (Validation 6)

```{r}
#| label: tbl-sens
#| tbl-cap: "Each row changes one modelling choice and re-runs the whole apportionment and calibration. Revenue is the calibrated central estimate."
variants <- tribble(
  ~Variant, ~args,
  "Central", list(),
  "Business allocator: excl. public administration", list(business_allocator = "lodes_ex_pubadmin"),
  "Business allocator: Econ Census payroll",         list(business_allocator = "ec_payroll"),
  "Delivered goods 0% to residence",                 list(delivered_to_residence = 0),
  "Delivered goods 100% to residence",               list(delivered_to_residence = 1),
  "Residence allocator: households not income",      list(resid_var = "hh"),
  "E-commerce: 15% of prone groups to residence",    list(ecommerce_frac = 0.15),
  "E-commerce: 30% of prone groups to residence",    list(ecommerce_frac = 0.30),
  "NAICS 4413 sourced to residence",                 list(mv_include_4413 = TRUE))

sens <- variants |>
  rowwise() |>
  mutate(B_pred_v = do.call(apportion_variant, c(ALB, args)),
         B_cal_v  = apply_calibration(fit, B_pred_v)$B_c_central,
         Rev      = rev_of(B_cal_v)) |>
  ungroup()

sens |>
  transmute(Variant,
            `Predicted base ($B)` = B_pred_v / 1e9,
            `Calibrated base ($B)` = B_cal_v / 1e9,
            `Revenue ($M)` = Rev / 1e6,
            `vs central` = Rev / REV - 1) |>
  gt() |>
  fmt_number(c(`Predicted base ($B)`, `Calibrated base ($B)`), decimals = 3) |>
  fmt_number(`Revenue ($M)`, decimals = 1) |>
  fmt_percent(`vs central`, decimals = 1) |>
  tab_options(table.font.size = pct(85))
```

The striking result is how little most of these matter. The delivered-goods
split — the plan's flagged 50/50 assumption — moves the answer by less than a
tenth of a percent, because the Economic Census store share and the ACS income
share for those groups happen to be nearly identical. The e-commerce assignment,
which the loss of NAICS 454 made unresolvable, moves it by well under 1% for the
same reason: reassigning receipts from a `r PC(0.2594)` allocator to a
`r PC(resid_share("01000","001","agginc"))` allocator barely changes anything.
**The unresolvable data problem turned out to be immaterial.**

What does matter is the **business allocator**. Excluding public administration
cuts the estimate by
`r sprintf("%.0f", abs(100*(sens$Rev[2]/REV-1)))`%. That is the real uncertainty
in this exercise, and it is a substantive question rather than a technical one:
how much taxable purchasing does state government actually do inside the city?
LODES counts the jobs; nothing public counts the purchases.

### Behavioural response, β

The add-on creates an 8.5% rate inside the city against 8.0% outside — a
6.25% increase in the tax, or 0.46% on a tax-inclusive price. Motor vehicles are
unaffected, since they are sourced to the purchaser's residence regardless of
where the dealer sits. Business purchases, utilities and services are largely
immobile. Only the store classes, about
`r PC(sum(appo$B_kg[appo$sourcing_class %in% c("store","delivered_split")])/B_k)`
of the base, are plausibly mobile at all, and only near the city line.

$\beta = 1.0$ is carried as central. A low case of
$\beta = `r sprintf("%.2f", BETA_LO)`$ — implying an elasticity of taxable sales
with respect to the tax-inclusive price of about 0.5 — gives
\$`r M(rev_of(est$B_c_central, BETA_LO))` M, a
`r sprintf("%.1f", 100*(1-BETA_LO))`% reduction. This is small relative to the
calibration band and does not change any conclusion.

The cross-border literature establishes that such responses exist and are
concentrated within a few miles of a rate discontinuity (Agrawal, 2015; Baker,
Johnson & Kueng, 2021). Neither paper studies a 0.5-point intra-county
differential in an upstate New York city, so **the specific low-case elasticity
above is illustrative** rather than transported from a published estimate, and
is labelled as such.

## 8. Timing and first-year cash

Local rate changes take effect only on 1 March, 1 June, 1 September or
1 December, and require 90 days' notice to the Commissioner (Tax Law §1210(d),
§1211). Vendors get transitional rules for straddling contracts and
subscriptions (Tax Bulletin ST-895). Distributions then lag liability by one to
two months, and the first two months of each quarter are paid as estimates and
trued up in the third.

Assuming a city fiscal year ending 31 December and an average cash lag of
1.5 months:

```{r}
#| label: tbl-timing
#| tbl-cap: "First-fiscal-year cash under the two most likely start dates, at the central annual estimate."
tibble(`Start date` = c("1 March", "1 December"),
       `Months of liability in first FY` = c(10, 1),
       `Months of cash in first FY` = c(8.5, 0),
       `First-year cash ($M)` = c(REV * 8.5/12, 0),
       `Steady state ($M)` = c(REV, REV) / 1e6) |>
  mutate(`First-year cash ($M)` = `First-year cash ($M)` / 1e6) |>
  gt() |>
  fmt_number(c(`First-year cash ($M)`, `Steady state ($M)`), decimals = 1) |>
  tab_options(table.font.size = pct(90))
```

A 1 March start captures roughly \$`r M(REV*8.5/12)` M of cash in the first
calendar year and reaches the full \$`r M(REV)` M in the second. A 1 December
start yields essentially nothing in the first calendar year — the single month of
December liability arrives as cash in mid-January. If the objective is revenue
in a particular budget year, the start date is worth as much as several of the
modelling choices in §7.

## 9. Validation summary

| # | Validation | Result |
|---|---|---|
| 1 | 4% × county base vs county distributions | Gap `r PC(min(-phi_tab$phi),1)` to `r PC(max(-phi_tab$phi),1)`; passes (plan threshold 10%) |
| 2 | Economic Census places vs county total | Places cover 99.2% of county retail receipts; residual 0.8% |
| 3 | Utilities: allocator vs measured school-district base | `r PC(work_share("3601000","36001"))` vs `r PC(albany_measured_utility_share())` — agree within `r sprintf("%.0f", abs(100*(work_share("3601000","36001")/albany_measured_utility_share()-1)))`% |
| 4 | Motor vehicles: DMV vs ACS income share | `r PC(mv_share("3601000","ALBANY"))` vs `r PC(resid_share("01000","001","agginc"))`; DMV lower, consistent with lower urban vehicle ownership |
| 5 | Calibration fit | β = `r sprintf("%.3f", fit$beta)` (se `r sprintf("%.3f", fit$se_beta)`), σ = `r sprintf("%.3f", fit$sigma)` — **exceeds the 0.25 threshold; band widened** |
| 6 | Sensitivity table | §7; business allocator dominates, everything else < 1% |
| 7 | Albany's R against the Route B distribution | `r sprintf("%.2f", R_alb)` against a range of `r sprintf("%.2f–%.2f", min(route_b$R), max(route_b$R))`, median `r sprintf("%.2f", median(route_b$R))` — inside |

## 10. What would move this estimate

Honest ranking of what the reader should worry about.

1. **State government purchasing.** `r PC(pubadmin_share("3601000","36001"))` of
   county public-administration employment is in the city, and dropping public
   administration from the allocator moves revenue by
   `r sprintf("%.0f", abs(100*(sens$Rev[2]/REV-1)))`%. No public source measures
   what state agencies actually buy, where, or how much of it is taxable. This is
   the largest real uncertainty and it is not reducible with public data.
2. **The calibration spread itself.** σ = `r sprintf("%.3f", fit$sigma)` says the
   method misses by a third on cities where the truth is known. Albany may be an
   easier case than Salamanca or Ogdensburg — it is large, and its allocators
   resolve at fine NAICS detail — but nothing in the data proves that.
3. **Economic Census vintage and basis.** EC 2022 is a single-year snapshot
   applied to `r min(ANALYSIS_FYS)-1`–`r max(ANALYSIS_FYS)` bases. Because the
   allocators are shares, level drift matters much less than composition drift.
   More importantly, EC receipts are *gross*, not taxable, and are recorded at
   the establishment's location; the assumption is that the taxable fraction of a
   NAICS group is similar inside and outside the city.
4. **Remote sales.** DTF does not publish taxable sales by delivery
   jurisdiction. The residence allocator for delivered goods is an assumption,
   though §7 shows the estimate barely responds to it.
5. **Ogdensburg's rate**, one of seventeen calibration observations, rests on
   footnote reasoning rather than a published rate table (see Phase 2). It is
   small and does not drive the fit.

## References

Agrawal, D. R. (2015). The tax gradient: Spatial aspects of fiscal competition.
*American Economic Journal: Economic Policy, 7*(2), 1–29.
<https://www.aeaweb.org/articles?id=10.1257/pol.20120360>

Baker, S. R., Johnson, S., & Kueng, L. (2021). Shopping for lower sales tax
rates. *American Economic Journal: Macroeconomics, 13*(3), 209–250.
<https://www.aeaweb.org/articles?id=10.1257/mac.20190026>

New York State Department of Taxation and Finance. (2026). *Taxable sales and purchases quarterly data: Beginning sales tax year 2013–2014* [Data set]. <https://data.ny.gov/Government-Finance/Taxable-Sales-And-Purchases-Quarterly-Data-Beginni/ny73-2j3u/about_data>

New York State Department of Taxation and Finance. (2025). *State and local sales tax distributions: Beginning fiscal years ended March 31, 1995* [Data set]. <https://data.ny.gov/Government-Finance/State-and-Local-Sales-Tax-Distributions-Beginning-/5g2s-tnb7>

New York State Department of Taxation and Finance. (2026). *AS001—Revenue distribution certification*. <https://www.tax.ny.gov/research/stats/statistics/sales_tax/government/as001.htm>

New York State Department of Taxation and Finance. (2025). *Publication 718: New York State sales and use tax rates by jurisdiction*. <https://www.tax.ny.gov/pdf/publications/sales/pub718.pdf>

New York State Department of Taxation and Finance. (2025). *Publication 718-A: Enactment and effective dates of sales and use tax rates*. <https://www.tax.ny.gov/pdf/publications/sales/pub718a.pdf>

New York State Department of Taxation and Finance. (2025). *Publication 718-C: Sales and use tax rates on clothing and footwear*. <https://www.tax.ny.gov/pdf/publications/sales/pub718c.pdf>

New York State Department of Taxation and Finance. (1990). *TSB-M-90(6)S: The Albany City School District imposes a 3% sales and use tax on utilities and utility services*. <https://tax.ny.gov/pdf/memos/sales/m90_6s.pdf>

New York State Department of Taxation and Finance. (2014). *Tax Bulletin ST-895: Transitional provisions for sales tax rate changes*. <https://www.tax.ny.gov/pubs_and_bulls/tg_bulletins/st/transitional_provisions.htm>

New York State Department of Motor Vehicles. (2026). *Vehicle, snowmobile, and boat registrations* [Data set]. Open NY. <https://data.ny.gov/Transportation/Vehicle-Snowmobile-and-Boat-Registrations/w4pv-hbkt>

N.Y. Tax Law §§ 1210, 1211, 1214, 1223, 1224.

Office of the New York State Comptroller. (2020). *Understanding local government sales tax in New York State: 2020 update*. <https://www.osc.ny.gov/files/local-government/publications/pdf/understanding-local-government-sales-tax-in-nys-2020-update.pdf>

U.S. Census Bureau. (2021). *2020 Census redistricting data (P.L. 94-171) summary file* [API dataset `dec/pl`]. <https://api.census.gov/data/2020/dec/pl.html>

U.S. Census Bureau. (2021). *2020 Census block assignment files*. <https://www2.census.gov/geo/docs/maps-data/data/baf2020/>

U.S. Census Bureau. (2024). *American Community Survey 5-year estimates: 2023* [API dataset `acs/acs5`]. <https://api.census.gov/data/2023/acs/acs5.html>

U.S. Census Bureau. (2025). *County Business Patterns: 2023* [API dataset `cbp`]. <https://api.census.gov/data/2023/cbp.html>

U.S. Census Bureau. (2025). *Economic Census: Summary statistics for the U.S., states, and selected geographies: 2022* [API dataset `ecnbasic`]. <https://api.census.gov/data/2022/ecnbasic.html>

U.S. Census Bureau. (n.d.). *2020 relationship files: ZCTA to place, ZCTA to block*. <https://www.census.gov/geographies/reference-files/time-series/geo/relationship-files.html>

U.S. Census Bureau, Longitudinal Employer-Household Dynamics. (2025). *LEHD Origin-Destination Employment Statistics (LODES) version 8: Workplace Area Characteristics, New York, 2023*. <https://lehd.ces.census.gov/data/>

---

*Every figure in this document is computed at render time from cached pulls in
`data/raw/` by `R/00_setup.R` … `R/05_calibrate.R`. Provenance for each raw file
— source URL, pull timestamp, SHA-256 — is in `data/raw/MANIFEST.csv`. The only
hand-entered inputs are the two crosswalk tables in `data/crosswalk/`, each
carrying a source URL per row. See `docs/data-layouts.md` for the verified schema
of every dataset and the ten places where it differed from what was assumed.*
