---
title: "Route B: the city share of the county sales tax base in New York's preempting cities"
subtitle: "A reduced-form benchmark for a City of Albany sales tax"
date: today
---

```{r setup}
#| include: false
library(here)
source(here("R", "05_calibrate.R"))
source(here("R", "02_pull_census_ec.R"))
suppressPackageStartupMessages({library(gt); library(ggplot2)})

panel   <- build_route_b()
route_b <- summarise_route_b(panel)
ident   <- check_identity(panel)
supp    <- suppression_report()
osw     <- oswego_evidence()
cw_all  <- read_crosswalk_cities()

## Albany reference quantities
alb_pop <- read_pop()
P_alb_c <- alb_pop$place$P_c[alb_pop$place$place_fips == "3601000"]
P_alb_k <- alb_pop$county$P_k[alb_pop$county$county_fips == "36001"]
alb_pop_share <- P_alb_c / P_alb_k

## the one directly observed Albany city/county base share (Phase 1, §4e)
dist <- read_distributions()
acsd <- mean(dist$amt[dist$fy %in% ROUTE_B_FYS &
                      dist$taxing_jurisdiction == "Albany City School District Consumer Utilities"])
base <- read_county_base()
alb_util <- base |>
  filter(jurisdiction == "ALBANY", fy %in% ROUTE_B_FYS,
         substr(naics_industry_group, 1, 2) == "22" | substr(naics_industry_group, 1, 3) == "517") |>
  group_by(fy) |> summarise(B = sum(B, na.rm = TRUE), .groups = "drop")
a_util <- (acsd / 0.03) / mean(alb_util$B)
R_util <- a_util / alb_pop_share

## 2022 Economic Census retail, Albany County (docs/data-layouts.md §5)
ec_p   <- ec_albany_places("44-45")
ec_k   <- ec_albany_county("44-45")
ec_alb <- ec_p$RCPTOT[grepl("^Albany city", ec_p$NAME)]
ec_col <- ec_p$RCPTOT[grepl("^Colonie town", ec_p$NAME)]
ec_share_city <- ec_alb / ec_k$RCPTOT
ec_resid <- 1 - sum(ec_p$RCPTOT, na.rm = TRUE) / ec_k$RCPTOT

## Albany's retail-only implied R, for reference on the plot
R_retail <- ec_share_city / alb_pop_share

## ACS aggregate household income share (the §4c delivered-goods allocator)
acs_p <- read_csv(here("data","raw","census_acs5_2023_place_ny.csv"),
                  col_types = cols(.default = col_character()))
acs_k <- read_csv(here("data","raw","census_acs5_2023_county_ny.csv"),
                  col_types = cols(.default = col_character()))
inc_share <- as.numeric(acs_p$B19025_001E[acs_p$place == "01000"]) /
             as.numeric(acs_k$B19025_001E[acs_k$county == "001"])

breakeven <- albany_breakeven()
fmt_pct <- function(x, d = 1) sprintf(paste0("%.", d, "f%%"), 100 * x)
```

## What this document does

The City of Albany does not levy a sales tax, so its sales tax base is not
observed. Eighteen New York cities outside New York City *do* levy one, and for
them the city base can be recovered from published distributions. This document
measures, for those cities, how concentrated the county sales tax base is inside
the city relative to population:

$$R_c \;=\; \frac{B_c / B_k}{P_c / P_k}$$

$R_c = 1$ means the city holds exactly its per-capita share of the county's
taxable sales. $R_c > 1$ means the city is a net retail centre for its county.
The distribution of $R_c$ across these cities is the empirical benchmark against
which any estimate of Albany's own base has to be judged, and it is the
prerequisite for the calibration in Phase 3.

Nothing here is an estimate of Albany's revenue. It is a benchmark.

## How the city base is recovered

Within county $k$ the county collects its full rate $r_k$ outside any preempting
city, and a reduced rate $r_k - p_c$ inside city $c$, while the city collects
$r_c$ on the same base. Collections $C$ are published; rates are verified from
Publication 718 and 718-A. So

$$B_c = \frac{C_c}{r_c}, \qquad C_k = r_k B_k - \sum_i p_i B_i$$

with $B_k$ — the whole-county base, including the preempting cities — taken
directly from the DTF taxable sales file.

Two things in that paragraph were assumptions in the project plan and are now
verified facts. First, `ny73-2j3u` reports the **whole county**, not the county
net of its preempting cities. Second, **$p_c = r_c$**: Publication 718 publishes
the combined rate both inside each preempting city and for the county
"– except" that city, and for every city but Yonkers the two are identical,
which forces the county's rate reduction to equal the city's rate exactly. The
project plan's assumption of a uniform $p_c = 1.5$ is wrong wherever
$r_c \neq 1.5$, which is nine of the eighteen cities.

Both facts are tested jointly below.

## Rates actually used

```{r}
#| label: tbl-rates
#| tbl-cap: "Verified rates. `r_k` is the county rate, `r_c` the city rate, `p_c` the county's rate reduction inside the city. Sources are Publication 718 (combined rates, effective 1 March 2025) and 718-A (enactment history); the full per-row citation is in `data/crosswalk/preempt_cities.csv`."
cw_all |>
  transmute(City = city, County = county,
            `r_k` = r_k, `r_c` = r_c, `p_c` = p_c,
            Used = ifelse(include, "yes", "excluded")) |>
  arrange(City) |>
  gt() |>
  fmt_number(c(r_k, r_c, p_c), decimals = 2) |>
  sub_missing(columns = p_c, missing_text = "—") |>
  tab_style(style = cell_fill(color = "#fdeaea"),
            locations = cells_body(rows = Used == "excluded")) |>
  tab_options(table.font.size = pct(90))
```

::: {.callout-warning}
## Publication 718-A's percentage column is not the city's imposed rate

This needs stating because it nearly derailed the numbers. Publication 718-A
lists, for six of these cities, a percentage that is **not** the rate the city
actually imposes: Gloversville and Johnstown 3 (actual 2.0), New Rochelle 3
(2.5), Norwich 3 (1.5), Saratoga Springs 3 (1.5). Taking that column at face
value doubles those cities' estimated base.

Three independent checks agree against it. Publication 718 gives the combined
rate inside each city, which fixes `county-in-city + r_c` exactly. Publication
718-A's own lettered footnotes give the county's rate inside the city
(footnote a "within city, county rate is 1½% (not counting any additional
rate)", c "1½%", t "2%", m "additional 1% not subject to preemption"), and
subtracting each from the combined rate reproduces the OSC value every time.
And the Office of the State Comptroller publishes the city rates directly,
sourced to DTF. Publication 718-A itself warns in its header that "the rates
indicated cannot be added to determine the combined state, county, and city tax
rate."

The rates above are therefore the OSC / footnote / Pub 718 set. The 718-A
column is retained in `data/crosswalk/preempt_cities.csv` as `r_c_pub718a_col`
so the alternative can be recomputed.

**Ogdensburg is the one weak row.** It repealed its tax in 2000 and re-imposed
it effective 1 March 2022, so it is absent from OSC's January 2020 table and its
rate rests on the footnote reasoning alone. If it is 3.0 rather than 1.5, its
$R$ falls from `r sprintf("%.2f", route_b$R[route_b$city=="Ogdensburg"])` to
about half that. It is one small observation and changes no conclusion here, but
it is the row to confirm with DTF.
:::

**Yonkers is the case that matters most.** Its combined rate is 8⅞% inside the
city against 8⅜% in the rest of Westchester. The city takes the county's full
4% and imposes 4.5% — a **0.5-point add-on above full preemption**. It is the
only published New York example of a city rate sitting above its county's, and
therefore the closest existing analogue to the policy this project is costing.
Its two distribution rows stand in an exact 2.000 ratio (3.0% regular plus 1.5%
special), so its base is the same whether computed from the total at 4.5% or the
regular row at 3.0% — the one city whose base is immune to the rate question
above.

## Validation: does the preemption identity hold?

If $p_i = r_i$ then the rates cancel in aggregate and
$\left(C_k + \sum_i C_i\right) / (r_k B_k)$ must equal 1. This tests the
identity and the whole-county reading of $B_k$ jointly, and — because the
observed collections appear on both sides — it is insensitive to the individual
city rates.

```{r}
#| label: tbl-identity
#| tbl-cap: "Total local collections divided by the county rate applied to the whole county base, three-year mean, FY2023–FY2025."
ident |>
  transmute(County = county,
            `County collections ($M)` = C_k / 1e6,
            `City collections ($M)`   = C_cities / 1e6,
            `County base ($B)`        = B_k / 1e9,
            `Ratio`                   = implied) |>
  gt() |>
  fmt_number(c(`County collections ($M)`, `City collections ($M)`), decimals = 1) |>
  fmt_number(`County base ($B)`, decimals = 2) |>
  fmt_number(Ratio, decimals = 3) |>
  tab_options(table.font.size = pct(90))
```

Every county lands between 0.995 and 1.027. The residual few tenths of a percent
are what one expects from the one-month offset between the sales tax year
(March–February) and the state fiscal year (April–March), from late and amended
returns, and from the administrative deduction. The identity holds.

::: {.callout-important}
## Oswego is excluded

Oswego is the one county where this fails, and it fails by too much to be noise.

```{r}
#| label: tbl-oswego
#| tbl-cap: "Oswego County: collections against 4% of the published county base."
osw |>
  filter(fy >= 2015) |>
  transmute(`Fiscal year` = fy,
            `County base ($B)` = B_k / 1e9,
            `County ($M)` = C_k / 1e6,
            `City ($M)` = C_c / 1e6,
            `County ÷ 4%·base` = county_only,
            `(County+City) ÷ 4%·base` = county_plus_city) |>
  gt() |>
  fmt_number(`County base ($B)`, decimals = 2) |>
  fmt_number(c(`County ($M)`, `City ($M)`), decimals = 1) |>
  fmt_number(c(`County ÷ 4%·base`, `(County+City) ÷ 4%·base`), decimals = 3) |>
  tab_options(table.font.size = pct(85))
```

Three published facts cannot all be true at once. Publication 718 gives 8%
combined both inside and outside the city, which implies the city preempts the
county's entire 4%. But Oswego County's collections alone equal essentially
100% of 4% × the county base in *every year from 2014 to 2025*, as though no
preemption occurred at all. And the city collects a further \$23 M on top.
Under any single interpretation of the statute one of those three has to give.

Publication 718-A flags the city with footnotes reading "county could not
preempt any of city's 3%" and "county may not preempt any of city's 4%", which
matches the collections but not the published combined rate. $p_c$ is not
identified, so per the project plan Oswego is reported here and dropped from
everything that follows. It is one observation of seventeen; excluding it does
not drive any conclusion below.
:::

## The result

```{r}
#| label: tbl-routeb
#| tbl-cap: "Observed city share of the county sales tax base, three-year mean of FY2023–FY2025. Population is the 2020 Census (P1_001N)."
route_b |>
  transmute(City = city, County = county,
            `City base ($M)`   = B_c / 1e6,
            `County base ($B)` = B_k / 1e9,
            `Base share s`     = s,
            `Pop. share`       = pop_share,
            `R` = R) |>
  gt() |>
  fmt_number(`City base ($M)`, decimals = 0) |>
  fmt_number(`County base ($B)`, decimals = 2) |>
  fmt_percent(c(`Base share s`, `Pop. share`), decimals = 1) |>
  fmt_number(R, decimals = 3) |>
  data_color(columns = R, palette = c("#c6dbef", "#f7f7f7", "#fdd0a2"),
             domain = c(0.3, 2.0)) |>
  tab_options(table.font.size = pct(90))
```

```{r}
#| label: fig-routeb
#| fig-cap: "Ratio of the city's share of its county's taxable sales base to its share of county population, three-year mean FY2023–FY2025. Oswego excluded. The dashed red line is Albany's break-even for preemption. The two dotted lines are Albany's only observable base shares: retail receipts from the 2022 Economic Census (orange) and utilities from the school district tax (blue). Albany's overall R is a base-weighted mix of those and the components Phase 3 has yet to allocate."
#| fig-height: 6.5
#| fig-width: 8
pd <- route_b |>
  mutate(city = factor(city, levels = rev(city[order(R)])),
         grp  = ifelse(county == "Westchester", "Westchester", "Rest of state"))

ggplot(pd, aes(R, city, colour = grp)) +
  geom_vline(xintercept = 1, colour = "grey75", linewidth = 0.4) +
  geom_vline(xintercept = breakeven, linetype = "dashed", colour = "#b2182b", linewidth = 0.55) +
  geom_vline(xintercept = R_util, linetype = "dotted", colour = "#2166ac", linewidth = 0.7) +
  geom_vline(xintercept = R_retail, linetype = "dotted", colour = "#e08214", linewidth = 0.7) +
  geom_segment(aes(x = 0, xend = R, yend = city), colour = "grey85", linewidth = 0.4) +
  geom_point(size = 3.1) +
  annotate("text", x = breakeven, y = 0.6, label = sprintf("Albany break-even %.3f", breakeven),
           hjust = -0.05, size = 3.1, colour = "#b2182b") +
  annotate("text", x = R_util, y = 17.5, label = sprintf("Albany utilities %.2f", R_util),
           hjust = 1.05, size = 3.1, colour = "#2166ac") +
  annotate("text", x = R_retail, y = 17.5, label = sprintf("Albany retail %.2f", R_retail),
           hjust = -0.05, size = 3.1, colour = "#e08214") +
  scale_colour_manual(values = c("Westchester" = "#762a83", "Rest of state" = "#1b7837"), name = NULL) +
  scale_x_continuous(breaks = seq(0, 2, 0.25), limits = c(0, 2.05), expand = c(0, 0)) +
  labs(x = expression(R[c]~"= (city base share) ÷ (city population share)"), y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "top")
```

### What the distribution says

```{r}
#| label: tbl-stats
tibble(Statistic = c("n cities", "Median", "Mean", "SD", "Minimum", "Maximum",
                     "Share above 1.0", "Share above Albany's break-even"),
       Value = c(sprintf("%d", nrow(route_b)),
                 sprintf("%.3f", median(route_b$R)), sprintf("%.3f", mean(route_b$R)),
                 sprintf("%.3f", sd(route_b$R)), sprintf("%.3f", min(route_b$R)),
                 sprintf("%.3f", max(route_b$R)),
                 sprintf("%d of %d", sum(route_b$R > 1), nrow(route_b)),
                 sprintf("%d of %d", sum(route_b$R > breakeven), nrow(route_b)))) |>
  gt() |> tab_options(table.font.size = pct(90), column_labels.hidden = TRUE)
```

The distribution **straddles 1.0 and is enormously wide**. The median is
`r sprintf("%.2f", median(route_b$R))` and
`r sum(route_b$R > 1)` of `r nrow(route_b)` cities hold more than their
per-capita share of the county base — but the range runs from
`r sprintf("%.2f", min(route_b$R))` to `r sprintf("%.2f", max(route_b$R))`, a
factor of four, with a standard deviation of
`r sprintf("%.2f", sd(route_b$R))` around a mean of
`r sprintf("%.2f", mean(route_b$R))`. Knowing that a New York city imposes a
sales tax tells you almost nothing about how concentrated its county's base is
inside it.

That spread is the single most important result for Phase 3. It means the
reduced form cannot pin Albany down on its own: an estimate of Albany's base
built by apportionment has to be defended on its construction, not by appeal to
where comparable cities sit, because "comparable cities" span a fourfold range.
It also sets a floor on how narrow any honest error band can be.

The spread is not random, though, and the pattern is legible. The cities at the
top are small cities that serve as the retail hub of a rural or small
metropolitan county — Oneida in Madison, Johnstown in Fulton, Saratoga Springs
in Saratoga, Ithaca in Tompkins, Olean in Cattaraugus — joined by White Plains,
a genuine regional commercial centre. The cities at the bottom are dense
residential cities inside counties whose retail has moved to suburban
corridors: Yonkers and Mount Vernon in Westchester, Salamanca in Cattaraugus,
Glens Falls in Warren, Utica in Oneida. Suburbanised retail pulls $R_c$ well
below 1; being the only town for thirty miles pushes it well above.

**Albany sits with the second group on the evidence available.** Albany County's
retail is concentrated along the Wolf Road corridor in Colonie and at Crossgates
in Guilderland, both outside the city. The 2022 Economic Census makes this
concrete: Albany city recorded
\\$`r sprintf("%.2f", ec_alb/1e6)` B of retail receipts against
\\$`r sprintf("%.2f", ec_col/1e6)` B in the balance of the town of Colonie
alone, and the city's `r fmt_pct(ec_share_city)` share of county retail
receipts is *below* its `r fmt_pct(alb_pop_share)` population share — a
retail-only $R$ of `r sprintf("%.2f", R_retail)`.

```{r}
#| label: tbl-eccov
#| tbl-cap: "2022 Economic Census retail trade (NAICS 44-45) receipts, Albany County places against the county total. Coverage check for the §4a store-based allocator."
ec_p |>
  transmute(Place = sub(", New York$", "", NAME), `Receipts ($000)` = RCPTOT) |>
  arrange(desc(`Receipts ($000)`)) |>
  bind_rows(tibble(Place = "— sum of places —",
                   `Receipts ($000)` = sum(ec_p$RCPTOT, na.rm = TRUE)),
            tibble(Place = "Albany County (published total)",
                   `Receipts ($000)` = ec_k$RCPTOT),
            tibble(Place = "residual (unincorporated / withheld)",
                   `Receipts ($000)` = ec_k$RCPTOT - sum(ec_p$RCPTOT, na.rm = TRUE))) |>
  gt() |>
  fmt_number(`Receipts ($000)`, decimals = 0) |>
  tab_style(style = cell_text(style = "italic"),
            locations = cells_body(rows = grepl("^—|^Albany County|^residual", Place))) |>
  tab_options(table.font.size = pct(85))
```

The places account for `r sprintf("%.1f%%", 100*(1-ec_resid))` of the county's
published retail receipts, so the store-based allocator has a solid geographic
foundation. Note that this is a *receipts* share, not a *taxable sales* share,
and receipts are recorded at the establishment's location.

### The two reference lines

The dashed line at `r sprintf("%.4f", breakeven)` is where preempting 1.5 points
would exactly replace what Albany now receives from the county's distribution.
`r sum(route_b$R > breakeven)` of `r nrow(route_b)` cities clear it. That
threshold depends on the 0.40 distribution factor carried in from the project
plan, which is **not yet verified** — confirming Albany County's actual
distribution formula is a Phase 3 task, and the line will move if the factor is
wrong.

The dotted line at `r sprintf("%.2f", R_util)` is the only directly observed
city/county base share Albany has. The Albany City School District levies a 3%
tax on utilities and telecommunications within the city; dividing its
collections by 0.03 gives a city base of
\$`r sprintf("%.0f", (acsd/0.03)/1e6)` M against
\$`r sprintf("%.0f", mean(alb_util$B)/1e6)` M of county NAICS 22 + 517 taxable
sales, so $a_\text{utilities}$ = `r fmt_pct(a_util)` against a population share
of `r fmt_pct(alb_pop_share)`.

That single point sits near the top of the distribution — but it should not be
read as an estimate of Albany's overall $R$, for two reasons. Utilities and
telecommunications are the *most* city-concentrated part of any tax base,
because commercial and institutional electricity, gas and telecom load
concentrates where the office buildings are, and in Albany's case that includes
the State Capitol complex. And the measurement is not clean: TSB-M-90(6)S
establishes that residential gas, electricity and steam are exempt from the
state tax and taxed by the county at a reduced 1% while the school district
taxes them at the full 3%, so numerator and denominator cover somewhat
different mixes.

### Where Albany's estimate will have to land

The two Albany data points available bracket the answer rather than pin it. The
store-based part of Albany's base looks like
$R \approx `r sprintf("%.2f", R_retail)`$; the utility part is measured at
$R \approx `r sprintf("%.2f", R_util)`$. Albany's overall $R$ is a
base-weighted mix of those and of the business-purchase and delivered-goods
components that Phase 3 has yet to allocate — the first of which, driven by
state government and downtown office employment, should be strongly
city-concentrated, and the second of which, driven by residence, should not be
(the city's aggregate household income share is only
`r fmt_pct(inc_share)`).

So the reduced form does not deliver Albany's answer. What it delivers is a
range to be judged against: an apportionment result landing anywhere in
`r sprintf("%.2f–%.2f", min(route_b$R), max(route_b$R))` is unremarkable, and
one landing outside it needs a specific defence. If Phase 3 returns an $R$ near
the top of that range it should be doubted, because Albany's largest observable
component — retail — is *below* parity, and no city in this sample reaches the
top of the range while ceding its county's retail to a suburban corridor the way
Albany does.

## Data quality note: cell suppression

DTF withholds individual county × NAICS cells for disclosure. There is no
jurisdiction total row in `ny73-2j3u`, so a county base is necessarily the sum
of its *published* groups.

```{r}
#| label: tbl-supp
#| tbl-cap: "Withheld county × NAICS cells, mean over FY2023–FY2025, for Albany and the analysis counties."
supp |>
  filter(jurisdiction %in% c("ALBANY", unique(panel$dtf_jurisdiction))) |>
  transmute(Jurisdiction = jurisdiction, `NAICS groups` = n_groups,
            `Withheld` = n_withheld, `% withheld` = pct_withheld,
            `Published base ($B)` = B_published / 1e9) |>
  arrange(desc(`% withheld`)) |>
  gt() |>
  fmt_number(c(`NAICS groups`, Withheld), decimals = 1) |>
  fmt_number(`% withheld`, decimals = 1) |>
  fmt_number(`Published base ($B)`, decimals = 2) |>
  tab_options(table.font.size = pct(85))
```

Between 3% and 10% of the 309 groups are withheld in any given county-year. The
withheld cells are small by construction — a cell is suppressed when few
taxpayers report it — and the identity check above bounds their total effect,
since it reconciles collections against the published base to within 3%. This
matters more in Phase 3, where the apportionment runs group by group, than it
does here.

## Sources

New York State Department of Taxation and Finance. (2026). *Taxable sales and purchases quarterly data: Beginning sales tax year 2013–2014* [Data set]. <https://data.ny.gov/Government-Finance/Taxable-Sales-And-Purchases-Quarterly-Data-Beginni/ny73-2j3u/about_data>

New York State Department of Taxation and Finance. (2025). *State and local sales tax distributions: Beginning fiscal years ended March 31, 1995* [Data set]. <https://data.ny.gov/Government-Finance/State-and-Local-Sales-Tax-Distributions-Beginning-/5g2s-tnb7>

New York State Department of Taxation and Finance. (2025). *Publication 718: New York State sales and use tax rates by jurisdiction*. <https://www.tax.ny.gov/pdf/publications/sales/pub718.pdf>

New York State Department of Taxation and Finance. (2025). *Publication 718-A: Enactment and effective dates of sales and use tax rates*. <https://www.tax.ny.gov/pdf/publications/sales/pub718a.pdf>

New York State Department of Taxation and Finance. (2025). *Publication 718-C: Sales and use tax rates on clothing and footwear*. <https://www.tax.ny.gov/pdf/publications/sales/pub718c.pdf>

New York State Department of Taxation and Finance. (1990). *TSB-M-90(6)S: The Albany City School District imposes a 3% sales and use tax on utilities and utility services*. <https://tax.ny.gov/pdf/memos/sales/m90_6s.pdf>

U.S. Census Bureau. (2021). *2020 Census redistricting data (P.L. 94-171) summary file* [API dataset `dec/pl`]. <https://api.census.gov/data/2020/dec/pl>

U.S. Census Bureau. (2025). *Economic Census: Summary statistics for the U.S., states, and selected geographies: 2022* [API dataset `ecnbasic`]. <https://api.census.gov/data/2022/ecnbasic.html>

---

*Every figure in this document is computed at render time from cached pulls in
`data/raw/` by `R/05_calibrate.R`. Provenance for each raw file — source URL,
pull timestamp, SHA-256 — is in `data/raw/MANIFEST.csv`. The rate table is the
only hand-entered input; each row carries its source URL in
`data/crosswalk/preempt_cities.csv`.*
